
---
class: inverse, middle, center

.center[
.font150[
**Backup**
]
]

<html>
  <div style='float:left'></div>
  <hr color='#005500' size=2px width=900px>
</html>

```{r tdc, child = "time-dependent-covariates.Rmd", eval = TRUE}
```

```{r input-censoring, child="censoring.Rmd", eval = TRUE}
```

---
# Piecewise exponential model (PEM)


$$h(t|\bfx_i) = h_0(t) \exp(\bfx_i^\top\bsbeta) = \exp(g_0(t) + \bfx_i^\top \bsbeta)$$
where the baseline hazard is written as $g_0(t) = \log h_0(t)$. <br>
We now want to estimate both $h_0(t)$ as well as the parameters
$\bsbeta$ via ML.

We decompose the time axis into $J$ intervals, using cut points $a_0, a_1, \ldots, a_J$:
$$(0=a_0,a_1],(a_1,a_2],\ldots,(a_{j-1},a_j],\ldots,(a_{J-1},a_J]$$
These interval cut points will often correspond to event and/or censoring times, but don't need to.

We assume piecewise constant hazards
$$h_0(t) = h_{0j} \quad \forall\, t \in I_j:=(a_{j-1},a_j]$$
with $\beta_{0j} = \log(h_{0j})$, this yields
$$h(t|\bfx_i) = h_{0j}\exp(\bfx_i^\top\bsbeta) = \exp(\beta_{0j} + \bfx_i^\top \bsbeta) = \exp(\eta_{ij})=: h_{ij} \quad \forall t \in I_j$$






