
---
class: inverse, middle, center

.center[
.font120[
**Competing Risks**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---
# Event History Analysis: Competing Risks Setting

.center[
<img src="figures/event-history-analysis.svg", width="700px">
]

- we are interested in the occurences of multiple, mutually exclusive events, e.g. patients can die from different causes or recover permanently.
- transient state $0$
- *absorbing* states $1,\ldots,K$


---
# Competing Risks - Example: Pneumonia on admission in intensive care unit patients

The example follows `r Citet(bib, "beyersmann_competing_2012")`: "Competing Risks and Multistate Models with R"

- 747 ICU patients
- endpoints: `status` = 0/1/2 for censoring, discharge from the ICU, and death
- event of interest: death (`status` = 2), competing event: discharge from the ICU (`status` = 1)
- variable of interest  `pneu`: A patient's pneumonia status at admission (0 = no pneumonia, 1 = pneumonia)

```{r, echo = TRUE}
data(sir.adm, package = "mvna")
sir.adm |> filter(id %in% c(41, 395, 17058)) |> knitr::kable()
```


---
# Competing Risks - Example: Pneumonia on admission in intensive care unit patients (2)

Data in PED format:

```{r echo = TRUE}
ped <- as_ped(sir.adm, Surv(time, status)~ pneu, combine = TRUE) |>
  mutate(cause = as.factor(cause))
ped |> filter(id == 41) |> group_by(id, cause) |> kable()
```

---
# Competing Risks - Example: Pneumonia on admission in intensive care unit patients (3)
```{r echo = TRUE, output.lines=c(8:13, 17:20)}
pam <- pamm(ped_status ~ s(tend, by = cause) + cause*pneu, data = ped)
summary(pam)
```

```{r echo = TRUE}
est <- c(Discharge = coef(pam)[3] |> unname(),       Death = sum(coef(pam)[3:4]))
se <-  c(Discharge = diag(vcov(pam))[3] |> unname(), Death = sum(vcov(pam)[3:4, 3:4])) |> sqrt()
cbind(estimate = est, lo = est - 2 * se, hi = est + 2 * se) |> exp() |> round(2)
```

---
# Competing Risks - Example: Pneumonia on admission in intensive care unit patients (4)

- pneumonia doesn't appear to have a clear effect on the death hazard
- this does *not* imply that pneumonia has no effect on mortality <br>
$\implies$ Calculate the CIF to check effect of pneumonia on probability of death

First, use `make_newdata` to create the data set to calculate the CIF

```{r, echo = TRUE}
ndf <- ped |> make_newdata(tend  = unique(tend), pneu  = unique(pneu), cause = unique(cause))
```

Second, use `add_cif` to add the CIF to the data set.
```{r echo = TRUE}
ndf <- ndf |>
  group_by(cause, pneu) |> # important!
  add_cif(pam) |> ungroup() |>
  mutate(
    cause = factor(cause, labels = c("Discharge", "Death")),
    pneu  = factor(pneu, labels = c("No Pneumonia", "Pneumonia"))
  )
```

---
# Competing Risks - Example: Pneumonia on admission in intensive care unit patients (5)

`add_cif` crated a new columns `cif`, `cif_lower`, and `cif_upper`, which can be visualized.

```{r pneu-viz, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "800px"}
# visualization
ggplot(ndf, aes(x = tend, y = cif)) + geom_line(aes(col = pneu)) +
  geom_ribbon(aes(ymin = cif_lower, ymax = cif_upper, fill = pneu), alpha = .3) +
  facet_wrap(~cause) + labs(y = "CIF", x = "time", color = "Pneumonia", fill = "Pneumonia")
```

---
# Competing Risks - Example: Pneumonia on admission in intensive care unit patients (6)

```{r pneu-viz, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "600px"}
```

- Pneumonia reduces the hazard for Discharge
$\implies$ subjects admitted with pneumonia stay on ICU longer

- While patients are in the ICU, they are exposed to a hazard for dying there
$\implies$ more patients with pneumonia die than patients without pneumonia
(even if *direct* effect of `pneu` on cause-specific hazard for death is unclear)
