
---
class: inverse, middle, center

.center[
.font120[
**Competing Risks**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---
# Event History Analysis

.center[
<img src="figures/event-history-analysis.svg", width="700px">
]

In the Competing Risks setting
- we are interested in multiple, mutually exclusive events, e.g. a patient may die from different causes
- $0$ transient state; $1,\ldots,K$ absorbing states


---
# Latent Failure Time Approach

Consider two event types $E \in \{1, 2\}$

$$T = \min(T_1,T_2), E = 1 \iff T_1 < T_2$$

- event type $E$ with associated event time $T_E$

- Only $(T, E)$ observable, joint distribution of $(T_1, T_2)$ is not identifiable

- assuming independence between $T_1$ and $T_2$ makes it identifiable (but checking the assumption is not possible using the observed data)

- interpretation of joint distribution rarely interesting

- in general, we prefer the **cause-specific hazard** model

---
# Cause-Specific Hazards

We define $T$ as the time to an event $E \in \{1,\ldots,K\}$.

The cause-specific hazards are
$$h_e(t) = \lim_{\Delta t \to 0} \frac{P(t \leq T \leq t + \Delta t, E = e\ |\ T \geq t)}{\Delta t}, \; e = 1, \dots, K.$$

From the cause specific hazards, we can derive the **cumulative incidence function (CIF)** (gives the probability of experiencing event $E = e$ before time $t$)
$$F_e(t) = P(T \leq t, E = e) = \int_0^t h_e(u) S(u)\ du,$$

using:
- cause specific cumulative hazard: $H_e(t) = \int^t_0 h_e(s)ds$
- all cause hazard: $h(t) = \sum_{e=1}^{K} h_e(t)$
- all cause cumulative hazard: $H(t) = \sum_{e=1}^{K} H_e(t)$
- all cause survival probability: $S(t) = \exp(-H(t))$


---
# Estimation
The distribution of the observed data is estimated by fitting cause-specific hazards models
$$h_e(t|\bfx_i) = h_{0e}(t) \exp(\bfx_i^\top \bsbeta_e)$$

To do so, we define two new status variables, such that

.pull-left[
.center[
data set

$$\begin{array}{llll}
T\quad & E\quad & \delta\quad & x\\ \hline
3 & 1 & 1 & x_1\\
4 & 2 & 1 & x_2\\
2 & 0 & 0 & x_3\\
\end{array}$$
]
]

.pull-right[
.center[
becomes

$$\begin{array}{llllll}
T & E & \delta & x & \delta_1 & \delta_2 \\ \hline
3 & 1 & 1 & x_1 & 1 & 0\\
4 & 2 & 1 & x_2 & 0 & 1\\
2 & 0 & 0 & x_3 & 0 & 0\\
\end{array}$$
]
]


where the cause-specific hazard estimates use hazard rate models with
event indicator $\delta_1$ for $e = 1$ and event indicator $\delta_2$ for $e = 2$.


---
# Cause-Specific PAMMs

- data transformation as usual with `as_ped`:
  + automatically recognizes if status variable has more then 2 unique values
  + creates one data set per non-censoring event
  + argument `combine` controls whether the separate data sets are retured as list or stacked into one data set with additional column `cause`

- fit cause-specific PAMMs to each data set (or jointly w/ interactions of `cause` variable with other predictors, including time)
$$h_e(t|\bfx) = \exp(g(t, e, \bfx))$$

- interpretation **conditional on *non-occurrence*** of all competing events <br>
  $\implies$ no (direct) inference on CIF, which is often more relevant
- however, we can calculate the CIF from the $h_e(t|\bfx), e=1,\ldots, K$


---
# Example: Pneumonia on admission in intensive care unit patients

The example follows `r Citet(bib, "beyersmann_competing_2012")`: "Competing Risks and Multistate Models with R"

- 747 ICU patients
- endpoints:
  + discharge from the ICU (`status` = 1): competing event
  + death (`status` = 2): event of interest
- variable of interest  `pneu`: A patient's pneumonia status at admission (0 = no pneumonia, 1 = pneumonia)

```{r, echo = TRUE}
data(sir.adm, package = "mvna")
sir.adm %>% filter(id %in% c(41, 395, 17058))
table(sir.adm$status)
sum(sir.adm$status != 0)
```


---
# Example: Pneumonia on admission in intensive care unit patients

Data in PED format:

```{r echo = TRUE}
ped <- as_ped(sir.adm, Surv(time, status)~ pneu, combine = TRUE) %>%
  mutate(cause = as.factor(cause))
dim(ped)
sum(ped$ped_status)
```

---
# Example: Pneumonia on admission in intensive care unit patients
```{r echo = TRUE}
ped %>%
  filter(id %in% c(41, 17058)) %>%
  group_by(id, cause) %>%
  slice(1, n()) |> kable()
```
---
# Example: Pneumonia on admission in intensive care unit patients
```{r echo = TRUE, output.lines=c(8:13, 17:20)}
pam <- pamm(ped_status ~ s(tend, by = cause) + cause*pneu, data = ped)
summary(pam)
```

```{r echo = TRUE}
est <- c(Discharge = coef(pam)[3] |> unname(),       Death = sum(coef(pam)[3:4]))
se <-  c(Discharge = diag(vcov(pam))[3] |> unname(), Death = sum(vcov(pam)[3:4, 3:4])) |> sqrt()  
cbind(estimate = est, hi = est + 2 * se, lo = est - 2 * se) |> exp() |> round(2)
```

---
# Example: Pneumonia on admission inintensive care unit patients

- pneumonia doesn't appear to have an effect on the death hazard
- this does *not* imply that pneumonia has no effect on mortality
- We want to calculate the CIF to check effect of pneumonia on probability of death

```{r, echo = TRUE}
ndf <- ped %>% make_newdata(tend  = unique(tend), pneu  = unique(pneu), cause = unique(cause))
ndf %>% group_by(cause) %>% slice(1:3) |> kable()
```

---
# Example: Pneumonia on admission inintensive care unit patients

```{r echo = TRUE}
ndf <- ndf %>%
  group_by(cause, pneu) %>% # important!
  add_cif(pam) %>% ungroup() %>%
  mutate(
    cause = factor(cause, labels = c("Discharge", "Death")),
    pneu  = factor(pneu, labels = c("No Pneumonia", "Pneumonia"))
  )
```

---
# Example: Pneumonia on admission inintensive care unit patients

```{r echo = TRUE, fig.width = 8, fig.height = 3, out.width = "600px"}
# visualization
ggplot(ndf, aes(x = tend, y = cif)) + geom_line(aes(col = pneu)) +
  geom_ribbon(aes(ymin = cif_lower, ymax = cif_upper, fill = pneu), alpha = .3) +
  facet_wrap(~cause) + labs(y = "CIF", x = "time", color = "Pneumonia", fill = "Pneumonia")
```

--
- Pneumonia reduces the hazard for Discharge $\rightarrow$ subjects admitted with pneumonia stay on ICU longer
- While they are in the ICU, they are exposed to an unchanged hazard for death
- As a consequence more patients with pneumonia die than patients without pneumonia
