
---
class: inverse, middle, center

.center[
.font150[
**1. Why is `pammtools` useful?** <br>
An overview of survival analysis results tackled in today's session
]
]

<html>
  <div style='float:left'></div>
  <hr color='#005500' size=2px width=900px>
</html>

```{r input-setup-pem, child="setup-pem.Rmd", eval = TRUE}
```

---
# With `pammtools`, you can calculate single event baseline hazards
**Example: Monoclonal gammopathy data `mgus` from the `survival` package**

```{r, est-pem-ex, echo=FALSE, fig.width = 6, fig.height=3, dependson=c("sim-wb", "ex-pem-1")}
ped2 <- as_ped(Surv(time, status)~., cut = seq(0,10, by=.1), data=sim_df_weibull)
pem2 <- glm(ped_status ~ interval, data = ped2, family=poisson(), offset=offset)
pem_haz_df2 <- int_info(ped2) %>%
  mutate(offset = intlen) %>%
  mutate(
    hazard = predict(object=pem2, ., type="response"),
    survival = exp(-cumsum(hazard * intlen)))
p_pem_haz2 <- p_pem_haz %+% pem_haz_df2 + ylim(c(0, .35))
```

```{r, est-pam-ex, echo=FALSE, fig.width = 6, fig.height=3, dependson=c("est-pem-ex")}
pam <- mgcv::gam(ped_status ~ s(tend), data = ped2, family=poisson(), offset=offset)
pem_haz_df2 <- pem_haz_df2 |> add_hazard(pam, overwrite = TRUE)
p_pam_haz <- p_pem_haz %+% pem_haz_df2 +
  geom_stepribbon(aes(ymin = ci_lower, ymax= ci_upper), alpha = .3) + ylim(c(0, .35))
```

```{r, echo=FALSE, fig.width = 6, fig.height = 6, dependson=c("est-pam-ex"), out.width = "800px"}
p_pam_haz_plot <- p_pam_haz + ggtitle("PAM")
```

```{r mgus-haz-df-ex, echo=FALSE}
data(mgus1, package="survival")
mgus1_clean <- mgus1 |> filter(event == "pcm")
mgus1_ped <- as_ped(data=mgus1_clean, 
                   Surv(stop, status) ~ .)
pam <- mgcv::gam(ped_status ~ s(tend), 
                 data= mgus1_ped, 
                 family=poisson(), 
                 offset = offset)
mgus_haz_df <- int_info(mgus1_ped) |> 
  mutate(offset = intlen) |> 
  add_hazard(pam) 
```

.pull-left[
**Task:** Calculate and plot the baseline hazard
- 241 observations
- `status`: 0/1 for censoring or interval ends in an event, here plasma cell malignancy
- `stop`: days from diagnosis to event
]

.pull-right[
```{r, echo=FALSE, eval=TRUE, dependson=c("mgus-haz-df-ex"), fig.width=5, fig.height=3, out.width="500px"}
ggplot(mgus_haz_df, aes(x = tend, y=hazard)) +
  geom_line() + ylim(c(0, 0.001)) + xlim(c(0,9000)) +
  geom_stepribbon(aes(ymin = ci_lower, ymax= ci_upper), alpha = .3) + xlab("time")
```
]


---
# With `pammtools`, you can do what `coxph` can do

**Example: Veterans' Administration Lung Cancer study** 

```{r vet-load-lin-ex, echo = FALSE}
veteran <- survival::veteran  |> filter(time < 400)  |> #reduce data for ex
  mutate(trt   = 1L * (trt == 2),
         prior = 1L * (prior == 10))

vet_cox_lin <- coxph(formula = Surv(time, status) ~ trt + prior + karno, 
                     data = veteran)
# data transformation
veteran_ped <- veteran |> as_ped(Surv(time, status)~., id = "id")
# fit model
vet_pam_lin <- pamm(ped_status ~ s(tend) + trt + prior + karno,
  data = veteran_ped)

```

.pull-left[
**Task:** Compare the linear effect of `trt`, `prior`, and `karno` between Cox and Pam. <br>
Randomised trial of survival times under two treatment regimens for lung cancer:
- endpoint: `status` (censoring = 0)
- variables of interest:  `karno` Karnovsky performance score (first: consider linear effect, later: time-varying effect), `trt` treatment (standard = 1, test = 0), `prior` prior therapy (no = 0, yes = 10)
]

.pull-right[
Compare Cox and PAM fits:
```{r vet-pam-coef-lin-ex, echo = FALSE}
cbind(
  pam = coef(vet_pam_lin)[2:4],
  cox = coef(vet_cox_lin)) |> kable()
```
]

---
# With `pammtools`, you can model non-linear effects

---
# With `pammtools`, you can model time-varying effects

**Example: Veterans' Administration Lung Cancer study** 

```{r vet-load-ex, echo = FALSE}
veteran <- survival::veteran  |> filter(time < 400)  |> #reduce data for ex
  mutate(trt   = 1L * (trt == 2),
         prior = 1L * (prior == 10))
# data transformation
veteran_ped <- veteran |> as_ped(Surv(time, status)~., id = "id") |>
    mutate(logt20 = log(tend + 20))
# fit model
vet_pam_logt <- pamm(ped_status ~ s(tend) + trt + prior + karno + karno:logt20,
  data = veteran_ped)
vet_pam_tve <- pamm(
  ped_status ~ s(tend) + trt + prior + s(tend, by = karno),
  data = veteran_ped)
```

.pull-left[
**Task (1):** Compare the linear effect and the smooth effect of the time and the Karnovsky performance score interaction <br>

Randomised trial of survival times under two treatment regimens for lung cancer
- endpoint: `status` (censoring = 0)
- variables of interest:  `karno` Karnovsky performance score (consider as TVE), `trt` treatment (standard = 1, test = 0), `prior` prior therapy (no = 0, yes = 10)
]
.pull-right[
```{r, echo =FALSE, fig.height = 3, fig.width = 5, out.width= "500px"}
term_df <- veteran_ped |> ped_info() |> add_term(vet_pam_tve, term = "karno") |>
  mutate_at(c("fit", "ci_lower", "ci_upper"), funs(. / .data$karno)) |>
  mutate(pam.fit = coef(vet_pam_logt)["karno"] + coef(vet_pam_logt)["karno:logt20"] * log(tend + 20))
ggplot(term_df, aes(x = tend, y = fit)) + geom_step(aes(col = "PAM spline")) +
    geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
    geom_step(aes(y = pam.fit, col = "PAM log(t)")) +
    scale_color_manual(name = "Model: ", values = c(Set1[1:2], "black")) +
    xlab("t") + ylab(expression(hat(beta)(t)))
```
]

---
# With `pammtools`, you can model time-varying effects
**Example: Veterans' Administration Lung Cancer study** 

.pull-left[
**Task (2):** Stratification, i.e. compare time-varying effects of factors between Cox and PAM <br>
- endpoint: `status` (censoring = 0)
- variables of interest:  `celltype` 1=squamous, 2=smallcell, 3=adeno, 4=large, interacting linearly (Cox) and smoothly (PAM) with `time`
]
.pull-right[

```{r, echo =FALSE, fig.height = 3, fig.width = 5, out.height=".4\textheight", warning = FALSE}
vet_cox_strata <- coxph(Surv(time, status) ~ strata(celltype),
                        data = veteran)
vet_pam_strata <- pamm(ped_status ~ s(tend, by = celltype) + celltype,
                       data = veteran_ped)

# extract fitted cumulative hazards:
pam_fit <- veteran_ped |>
  ped_info() |> slice(rep(1:n(), each = 4)) |>
  mutate(celltype = rep(levels(celltype), times = n()/4)) |>
  group_by(celltype) |> arrange(tend)  |>  #!!
  add_cumu_hazard(vet_pam_strata)
cox_fit <- vet_cox_strata |> basehaz()
```
```{r gg_vet_strata-ex, echo =FALSE, fig.height = 5, fig.width = 7, out.width  = "500px",}
ggplot(pam_fit, aes(x = tend, y = cumu_hazard)) + geom_step(aes(col = celltype)) +
    geom_stepribbon(aes(ymin = cumu_lower, ymax = cumu_upper, fill = celltype), alpha = 0.1) +
    geom_line(data = cox_fit, aes(x = time, y = hazard, col = strata), linetype = 4, size = 1) +
    ylab(expression(hat(H)(t))) + xlab("t") + coord_cartesian(ylim = c(0, 10))
```
]

---
# With `pammtools`, you can model competing risks

**Example: Pneumonia on admission in intensive care unit patients** <br>
The example follows `r Citet(bib, "beyersmann_competing_2012")`: "Competing Risks and Multistate Models with R"

```{r echo = FALSE}
data(sir.adm, package = "mvna")
ped <- as_ped(sir.adm, Surv(time, status)~ pneu, combine = TRUE) |>
  mutate(cause = as.factor(cause))
pam <- pamm(ped_status ~ s(tend, by = cause) + cause*pneu, data = ped)
ndf <- ped |> make_newdata(tend  = unique(tend), pneu  = unique(pneu), cause = unique(cause))
ndf <- ndf |>
  group_by(cause, pneu) |> # important!
  add_cif(pam) |> ungroup() |>
  mutate(
    cause = factor(cause, labels = c("Discharge", "Death")),
    pneu  = factor(pneu, labels = c("No Pneumonia", "Pneumonia"))
  )
```

.pull-left[
**Task:** Visualize the cause-specific transition hazards of "Discharge" and "Death"
- 747 ICU patients
- endpoints: `status` = 0/1/2 for censoring, discharge from the ICU, and death
- event of interest: death (`status` = 2), competing event: discharge from the ICU (`status` = 1)
- variable of interest  `pneu`: A patient's pneumonia status at admission (0 = no pneumonia, 1 = pneumonia)
]

.pull-right[
```{r, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "500px"}
# visualization
ggplot(ndf, aes(x = tend, y = cif)) + geom_line(aes(col = pneu)) +
  geom_ribbon(aes(ymin = cif_lower, ymax = cif_upper, fill = pneu), alpha = .3) +
  facet_wrap(~cause) + labs(y = "CIF", x = "time", color = "Pneumonia", fill = "Pneumonia")
```
]

---
# With `pammtools`, you can model multi-state settings
**Example: Abnormal prothrombin levels in liver cirrhosis**
.pull-left[
**Task:** Visualize treatment effect on transition probabilities into abnormal prothrombin level (and back) and death state
- 488 liver cirrhosis patients
- endpoints: `status` = 0/1 for censoring, event
- events of interest: `status` = 1 and `to` = 2/3 for abnormal prothrombin level, death
- variable of interest  `treat`: A patient's treatment (Placebo, Prednisone)
]

```{r, echo = FALSE}
data(prothr, package = "mstate")
prothr <- prothr %>% 
  rename(tstart = Tstart, tstop = Tstop) %>%
  filter(tstart != tstop)
# load function as it is not yet included in the package
devtools::load_all("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools")
#source("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools/R/add-functions.R")
source("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools/tmp/add_transition_probabilities.R")

library(tidyverse)

my.prothr <- prothr |> filter(status == 1) |> add_counterfactual_transitions() # add possible transitions

ped <- as_ped_multistate(
  data       = my.prothr,
  formula    = Surv(tstart, tstop, status)~ .,
  transition = "transition",
  id         = "id",
  timescale  = "calendar", 
  tdc_specials="concurrent"
)

ped$transition <- as.factor(ped$transition)
ped$treat <- as.factor(ped$treat)

pam <- pamm(ped_status ~ s(tend, by=transition) + transition * treat, data = ped)
ndf <- make_newdata(ped
        , tend  = unique(tend), treat  = unique(treat), transition = unique(transition))
ndf <- ndf |>
  group_by(treat, transition) |> # important!
  add_cumu_hazard(pam) 
old_groups <- dplyr::groups(ndf) # save old groups before ungrouping data
ndf_ungr <- ndf |> ungroup(transition) # probabilities differ for each transition
ndf <- group_split(ndf_ungr) |> # pipeline to calculate transition probability based on cumu hazards
  map(ndf_ungr, .f = ~ group_by(.x, transition)) |> 
  map(ndf_ungr, .f = ~ add_trans_prob(.x)) |>
  map(ndf_ungr, .f = ~ group_by(.x, !!!old_groups)) |>
  bind_rows()
```

.pull-right[
```{r prothr-viz-ex, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "500px"}
# visualization
ggplot(ndf, aes(x=tend, y=trans_prob)) + geom_line(aes(col=treat)) + facet_wrap(~transition) + 
  labs(y = "Transition Probability", x = "time", color = "Treatment", fill= "Treatment")
```
]