
---
class: inverse, middle, center

.font150[
**Part IV: Event History Analysis**
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>



---
class: inverse, middle, center

.font120[
**Left truncation**
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>



---
# Left Truncation

Only individuals that experienced an event *after* $y_i$ are observed, i.e., $t_i \geq y_i$.

Likelihood contribution:
$$L_i(\bstheta) = P(T=t_i | T \geq y_i| \bstheta) = \frac{f(t_i| \bstheta)}{S(y_i| \bstheta)}$$

For (non-parametric) hazard based methods: <br>
We can deal with left truncation, by only including subject in the risk after their left-truncation time

**Example: Infant Mortality** (1821–1894):
- Infants were included in the study when their mother died (left truncation event)
- Two infants of same age whose mothers were still alive were matched into the study
- All infants still alive were censored at the end of the study (365 days)
- Infants that die before their mother never enter the study
- Hypothesis: infants whose mothers are still alive survive longer


---
# Example: Infant Mortality


```{r, echo = TRUE}
data("infants", package = "eha")
infants$id <- seq_len(nrow(infants))
head(infants)
xtabs(~event + mother, data = infants)
summary(infants$enter)
summary(infants$exit)
```

---
# Left-truncation with PAMMs

- Data transformation:
  + as usual with `as_ped`, but specify left-truncation time and event time in formula
  + in each interval, only subjects that were at risks at that time, e.g.
    * infant enters study at day 5, when mother dies
    * not included in intervals $(0, 1], \ldots, (4, 5]$
    * included in intervals $(5, 6]$ ...
- Modell fit:
  + as usual based on properly transformed data
  + hazard can be estimated unbiasedly (assuming left-truncation time and event time are independent)


---
# Example: Infant Mortality (2)

```{r, echo = TRUE}
infants[1, c("id", "enter", "exit", "event", "mother", "age"), ]
ped_infants <- infants |> as_ped(Surv(exit, event) ~ .)
ped_infants |> filter(id == 1) |> select(tstart, tend, interval, ped_status, mother, age) |> slice(1, n())
ped_infants_lt <- infants |> as_ped(Surv(enter, exit, event) ~ .)
ped_infants_lt |>filter(id == 1) |> select(tstart, tend, interval, ped_status, mother, age)  |> slice(1, n())
```

---
# Example: Infant Mortality (3)

"Naive" analysis ignoring left-truncation:

```{r echo = TRUE, output.lines = c(8:14, 18:20)}
pam_infants <- pamm(
  ped_status ~ s(tend) +  mother + age + sex + parish + civst + ses,
  data = ped_infants)
summary(pam_infants) # truncated output
```

```{r echo = TRUE}
exp(coef(pam_infants)[2])
```

---
# Example: Infant Mortality (4)
Analysis that takes into account left-truncation:

.font90[
```{r echo = TRUE, output.lines = c(8:14, 18:20)}
pam_infants_lt <- pamm(
  ped_status ~ s(tend) +  mother + age + sex + parish + civst + ses,
  data = ped_infants_lt)
summary(pam_infants_lt) # truncated output
```
]

```{r echo = TRUE}
exp(coef(pam_infants_lt)[2])
```


---
# Left-truncation in Event History Analysis

- So far we only considered transitions from the initial status (e.g., alive) to a single event of interest (e.g., death)
- However, methods for Lifetime Data Analysis can also be applied to settings with
multiple states and transitions between them
- This more general setting is often refered to as *Event History Analysis*
.center[
<img src="figures/event-history-analysis.svg", width="700px">
]


---
# Left-truncation in Event History Analysis
.center[
<img src="figures/event-history-analysis.svg", width="600px">
]
- Left-truncation can occur in each of the "0" states
- In addition, left truncation is relevant for each (secondary) transition in a multi-state model (MSM), e.g.,
  + When estimating the hazard for transition "1 $\rightarrow$ 2", subjects arrive in state "1" at different time points
  + Similarly, for transitions "2 $\ra$ 1" and "2 $\ra$ 3"
  + State "3" is a competing risk w.r.t. state "1"
