

---
class: inverse, middle, center

.center[
.font150[
**What exactly do I need to do to get results?** <br>
An insight into the nitty-gritty details of producing survival analysis results with `pammtools` in R
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---
class: inverse, middle, center

.center[
.font120[
**Basic Scenarios**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>

---
# Linear Effect - Example: Veterans' Administration Lung Cancer study

Randomised trial of survival times under two treatment regimens for lung cancer:
- endpoint: `status` (censoring = 0)
- variables of interest:  `karno` Karnovsky performance score (first: consider linear effect, later: time-varying effect), `trt` treatment (standard = 1, test = 0), `prior` prior therapy (no = 0, yes = 10)
```{r vet-load-lin, echo = TRUE}
veteran <- survival::veteran  |> filter(time < 400)  |> #reduce data for ex
  mutate(trt   = 1L * (trt == 2),
         prior = 1L * (prior == 10))
rbind(head(survival::veteran, n = 2), tail(survival::veteran, n = 2))[, c(1:5, 8)] |> knitr::kable()
```

---
# Linear Effect - Example: Veterans' Administration Lung Cancer study

We start comparing the linear effect of `trt`, `prior`, and `karno` between Cox and Pam.

```{r vet-cox-lin, echo =TRUE, output.lines=c(6:9, 13:16)}
vet_cox_lin <- coxph(formula = Surv(time, status) ~ trt + prior + karno, 
                     data = veteran)
summary(vet_cox_lin)
```
---
# Linear Effect - Example: Veterans' Administration Lung Cancer study

In `pammtools`, we again create the PED. Then use the same pipeline as for Cox
```{r vet-pam-lin, echo =TRUE, output.lines=c(8:12, 17:19)}
# data transformation
veteran_ped <- veteran |> as_ped(Surv(time, status)~., id = "id")
# fit model
vet_pam_lin <- pamm(ped_status ~ s(tend) + trt + prior + karno,
  data = veteran_ped)
summary(vet_pam_lin)
```
---
# Linear Effect - Example: Veterans' Administration Lung Cancer study

Compare fits:
```{r vet-pam-coef-lin, echo = TRUE}
cbind(
  pam = coef(vet_pam_lin)[2:4],
  cox = coef(vet_cox_lin)) |> kable()
```