

---
class: inverse, middle, center

.center[
.font150[
**3. What exactly do I need to do to get results?** <br>
An insight into the nitty-gritty details of producing survival analysis results with `pammtools` in R
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---
class: inverse, middle, center

.center[
.font120[
**3.1 Basic Scenarios**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>

---
# Linear Effect - Example: Veterans' Administration Lung Cancer study

Randomised trial of survival times under two treatment regimens for lung cancer:
- endpoint: `status` (censoring = 0)
- variables of interest:
    - `karno` Karnovsky performance score (first: consider linear effect, later: time-varying effect),
    - `trt` treatment (standard = 0, test = 1),
    - `prior` prior therapy (no = 0, yes = 1)

```{r vet-load-lin, echo = FALSE}
veteran <- survival::veteran  |> filter(time < 400)  |> #reduce data for ex
  mutate(trt   = 1L * (trt == 2),
         prior = 1L * (prior == 10))
rbind(head(survival::veteran, n = 2), tail(survival::veteran, n = 2))[, c(1:5, 8)] |> knitr::kable()
```

---
# Linear Effect - Example: Veterans' Administration Lung Cancer study

We start comparing the linear effect of `trt`, `prior`, and `karno` between Cox and PAM.

.pull-left[
Cox
```{r vet-cox-lin, echo =TRUE}
vet_cox_lin <- coxph(
  formula = Surv(time,status)~trt+prior+karno,
  data = veteran)
```
]

.pull-right[
PAM
```{r vet-pam-lin, echo =TRUE}
# data transformation
vet_ped <- veteran |> as_ped(Surv(time,status)~.)
# fit model
vet_pam_lin <- pamm(
  formula = ped_status~s(tend)+trt+prior+karno,
  data = vet_ped)
```
]

```{r vet-pam-coef-lin, echo = FALSE}
cbind(
  cox = coef(vet_cox_lin),
  pam = coef(vet_pam_lin)[2:4]) |> kable()
```
