
---
class: inverse, middle, center

.center[
.font120[
**Multi-State Models**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---
# Event History Analysis: Multi-State Setting

.center[
<img src="figures/event-history-analysis.svg", width="700px">
]

- we are interested in the occurences of multiple, mutually exclusive events with in-between- and back-transitions, e.g. patients can transition between increasing severities of illnesses or recover permanently.
- transient state $0$
- *absorbing* state $3$


---
# Cause-Specific PAMMs

- data transformation as usual with `as_ped` / `as_ped_multistate`:
  + automatically recognizes if status variable has more than two unique values
  + automatically recognizes if id variable occurs multiple times in the data set
  + creates one data set per non-censoring event or stack them with an additional column `transition`

- fit cause-specific PAMMs jointly w/ interactions of `transition` variable with other predictors, including time, and shared effects for others
$$h_{k,e}(t|\bfx) = \exp(g(t, k, e, \bfx))$$

- transition probability depends on all competing events and current state at time $t$
- however, transition probabilities are easy to calculate from the cumulative hazards

---
# Example: Abnormal prothrombin levels in liver cirrhosis

- 488 liver cirrhosis patients
- endpoints: `status` = 0/1 for censoring, event
- events of interest: `status` = 1 and `to` = 2/3 for abnormal prothrombin level, death
- variable of interest  `treat`: A patient's treatment (Placebo, Prednisone)

```{r, echo = TRUE}
data(prothr, package = "mstate")
prothr |> filter(id == 46) |> knitr::kable() # example patients
```


---
# Example: Abnormal prothrombin levels in liver cirrhosis

```{r, echo = FALSE}
prothr <- prothr %>% 
  rename(tstart = Tstart, tstop = Tstop) %>%
  filter(tstart != tstop)
```

First, use `add_counterfactual_transitions` to add all possible transitions at each point in time:
```{r lib-ms-pammtools, echo = FALSE}
# load function as it is not yet included in the package
devtools::load_all("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools")
source("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools/tmp/add_transition_probabilities.R")
library(tidyverse)
```
```{r, echo = TRUE, dependson=c("lib-ms-pammtools")}
my.prothr <- prothr |> filter(status == 1) |> add_counterfactual_transitions() # add possible transitions
```

Second, use `as_ped_multistate` to create a PED object:
```{r echo = TRUE}
ped <- as_ped_multistate(
  data       = my.prothr,
  formula    = Surv(tstart, tstop, status)~ .,
  transition = "transition",
  id         = "id",
  timescale  = "calendar", 
  tdc_specials="concurrent"
)
```

---
# Example: Abnormal prothrombin levels in liver cirrhosis

Count events for each transition and treatment:
```{r echo = TRUE}
ped |> group_by(transition, treat) |> summarize(n_cases = sum(ped_status)) # observed events
sum(ped$ped_status) # total number of observed events
```


---
# Example: Abnormal prothrombin levels in liver cirrhosis
```{r, echo = FALSE}
ped$transition <- as.factor(ped$transition)
ped$treat <- as.factor(ped$treat)
```

```{r echo = TRUE, output.lines=c(8:17, 21:26)}
pam <- pamm(ped_status ~ s(tend, by=transition) + transition * treat, data = ped)
summary(pam)
```
---
# Example: Abnormal prothrombin levels in liver cirrhosis

First, use `make_newdata` to create the data set for prediction:
```{r, echo = FALSE}
# load function as it is not yet included in the package
devtools::load_all("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools")
```
```{r, echo = TRUE}
ndf <- make_newdata(ped
        , tend  = unique(tend), treat  = unique(treat), transition = unique(transition))
```

Second, use `add_cumu_hazard` to calculate the cumulative hazards
```{r echo = TRUE}
ndf <- ndf |>
  group_by(treat, transition) |> # important!
  add_cumu_hazard(pam) 
```

Third, use `add_trans_prob` to calculate the transition probabilities including back-transitions
```{r echo = TRUE, dependson=c("lib-ms-pammtools")}
old_groups <- dplyr::groups(ndf) # save old groups before ungrouping data
ndf_ungr <- ndf |> ungroup(transition) # probabilities differ for each transition
ndf <- group_split(ndf_ungr) |> # pipeline to calculate transition probability based on cumu hazards
  map(ndf_ungr, .f = ~ group_by(.x, transition)) |> 
  map(ndf_ungr, .f = ~ add_trans_prob(.x)) |>
  map(ndf_ungr, .f = ~ group_by(.x, !!!old_groups)) |>
  bind_rows()
```

---
# Example: Abnormal prothrombin levels in liver cirrhosis

`add_trans_prob` crated a new column `trans_prob`, which can be visualized.

```{r prothr-viz, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "800px"}
# visualization
ggplot(ndf, aes(x=tend, y=trans_prob)) + geom_line(aes(col=treat)) + facet_wrap(~transition) + 
  labs(y = "Transition Probability", x = "time", color = "Treatment", fill= "Treatment")
```

---
# Example: Abnormal prothrombin levels in liver cirrhosis

```{r prothr-viz, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "600px"}
```

- Prednisone treatment reduces the transition probability into the death state
and into abnormal prothrombin levels
- Prednisone treatment increases the probability to exit the state of abnormal prothrombin levels
$\implies$ subjects treated with Prednisone stay healthy longer

