

# Example: Chronic Granulotamous Disease (CGD)

- RCT for effect of interferon gamma (treatment) on patients with CGD
- CGD is an inherited disorder that manifests in recurrent infections (with skin leasons)

```{r, echo = TRUE}
data("cgd", package = "survival")
cgd |> filter(id == 1) |> select(id, treat, tstart, enum, tstop, status) |>
  mutate(gap = tstop - tstart)
table(cgd$enum)
```


---
# Example: Chronic Granulotamous Disease (CGD)

Transformation to PED format:

```{r echo = TRUE}
ped_cgd <- cgd |>
  mutate(id = factor(id)) |> # important for mgcv::gam
  as_ped(
    formula = Surv(tstart, tstop, status) ~ treat,
    id = "id",
    transition = "enum",
    timescale = "gap") |>
  mutate(enum_strata = factor(ifelse(enum > 2, "3+", enum)))
ped_cgd |> filter(id == 1) |> group_by(enum) |> slice(1, n())
```

---
# Example: Chronic Granulotamous Disease (CGD)

Fitting the model $\lambda(t|b_i) = \exp(\beta_0 + f_0(t) + \beta_1 x_{treat} + \ b_i),$ with
- $\beta_0 + f_0(t)$: baseline hazard
- $\beta_1$: treatment effect
- $b_i \sim N(0, \sigma_{b}^2)$: frailty/random intercept


```{r echo = TRUE, output.lines = c(8:11, 15:18)}
pam_cgd <- pamm(ped_status ~ s(tend) + treat + s(id, bs = "re"), data = ped_cgd)
summary(pam_cgd)
```

---
# Example: Chronic Granulotamous Disease (CGD)
```{r echo = TRUE, fig.width = 5, out.width = "400px", fig.keep = "last"}
qq_cgd <- gg_re(pam_cgd)
qq_cgd
```

---
# Example: Chronic Granulotamous Disease (CGD)
Fitting the model $\lambda(t|k, b_i) = \exp(\beta_{0k} + f_{0k}(t) + \beta_1 x_{treat} + \ b_i),$ with
- $\beta_{0k} + f_{0k}$: recurrence specific log baseline hazard

```{r echo = TRUE, output.lines = c(9:15, 19:25)}
pam_cgd2 <- pamm(ped_status ~ enum_strata + s(tend, by = enum_strata) + treat + s(id, bs = "re"), data = ped_cgd)
summary(pam_cgd2)
```

---
# Example: Chronic Granulotamous Disease (CGD)
Extract predicted survival probabilities per group:

```{r, echo = TRUE}
ndf_cgd <- ped_cgd |>
  make_newdata(tend = unique(tend), enum_strata = unique(enum_strata)) |>
  add_hazard(pam_cgd2) |>
  group_by(enum_strata) |> # !!! very important
  add_surv_prob(pam_cgd2)
ndf_cgd |> select(enum_strata, tend, surv_lower, surv_prob, surv_upper) |>
  group_by(enum_strata) |> slice(1, n())
```

---
# Example: Chronic Granulotamous Disease (CGD)

```{r echo = TRUE, fig.width = 7, out.width = "600px"}
ggplot(ndf_cgd, aes(x = tend, y = surv_prob)) +
  geom_surv(aes(col = enum_strata))
```
