

---
class: inverse, middle, center

.font150[
**How can `pammtools` help me analyze survival data?** <br>
A quick overview to get from survival data to interpretable results
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900px>
</html>

---
# `pammtools` exploits the benefits of piecewise exponential additive mixed models (PAMMs)

**Why exponential additive mixed model?**
- Same "hazard structure" as in Cox PH Models $implies$ exponential
- BUT can make use of GAMM advantages, e.g. smooth effects ($f$) dependent on $x$ and $t$, mixed effects $b$ for clusters $\ell$
$$\log(h_{k,e}(t|\mathbf{x}_{i, k, e}, \ell_i)) = \beta_{0,k,e} + f_{0,k,e}(t) + \sum_{p=1}^P f_p(x_{i,k,e,p}, t) + b_{\ell_i}$$
where the baseline hazard is written as $\beta_{0,k,e} + f_{0,k,e}(t)$. <br>
We now want to estimate both $h_0(t)$ as well as the parameters
$\bsbeta$ via ML.

**Why piecewise?** 
- Decompose the time axis into intervals
- Assume piecewise constant hazards in these intervals

**What data do we need for that?**


---
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped.svg", width = "400px" align="middle" >
</div>
</div>


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped1.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped2.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval

---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped3.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval
- Calculate differences of event times $t_i$ and interval endpoints $a_j$


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped4.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval
- Calculate differences of event times $t_i$ and interval endpoints $a_j$
- Define the "new" time points to the interval endpoints


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval
- Calculate differences of event times $t_i$ and interval endpoints $a_j$
- Define the "new" time points to the interval endpoints

Note: Often interval endpoints are event times.

---
# Use `pammtools` to get from survival data to illustrations
```{r, est-pem, echo=FALSE, fig.width = 6, fig.height=3, dependson=c("sim-wb", "ex-pem-1")}
ped2 <- as_ped(Surv(time, status)~., cut = seq(0,10, by=.1), data=sim_df_weibull)
pem2 <- glm(ped_status ~ interval, data = ped2, family=poisson(), offset=offset)
pem_haz_df2 <- int_info(ped2) %>%
  mutate(offset = intlen) %>%
  mutate(
    hazard = predict(object=pem2, ., type="response"),
    survival = exp(-cumsum(hazard * intlen)))
p_pem_haz2 <- p_pem_haz %+% pem_haz_df2 + ylim(c(0, .35))
```

```{r, est-pam, echo=FALSE, fig.width = 6, fig.height=3, dependson=c("est-pem")}
pam <- mgcv::gam(ped_status ~ s(tend), data = ped2, family=poisson(), offset=offset)
pem_haz_df2 <- pem_haz_df2 |> add_hazard(pam, overwrite = TRUE)
p_pam_haz <- p_pem_haz %+% pem_haz_df2 +
  geom_stepribbon(aes(ymin = ci_lower, ymax= ci_upper), alpha = .3) + ylim(c(0, .35))
```

```{r, echo=FALSE, fig.width = 6, fig.height = 6, dependson=c("est-pam"), out.width = "800px"}
p_pam_haz_plot <- p_pam_haz + ggtitle("PAM")
```

```{r mgus-haz-df, echo=FALSE}
data(mgus1, package="survival")
mgus1_clean <- mgus1 |> filter(event == "pcm")
mgus1_ped <- as_ped(data=mgus1_clean, 
                   Surv(stop, status) ~ .)
pam <- mgcv::gam(ped_status ~ s(tend), 
                 data= mgus1_ped, 
                 family=poisson(), 
                 offset = offset)
mgus_haz_df <- int_info(mgus1_ped) |> 
  mutate(offset = intlen) |> 
  add_hazard(pam) 
```

.pull-left[
Example (Baseline): Monoclonal gammopathy data `mgus` from the `survival` package
- 241 observations
- `status`: 0/1 for censoring or interval ends in an event, here plasma cell malignancy
- `stop`: days from diagnosis to event

```{r, echo=TRUE, eval=FALSE}
data(mgus1, package="survival")
mgus1_clean <- mgus1 |> filter(event == "pcm")
mgus1_ped <- as_ped(data=mgus1_clean, 
                   Surv(stop, status) ~ .)
pam <- mgcv::gam(ped_status ~ s(tend), 
                 data= mgus1_ped, 
                 family=poisson(), 
                 offset = offset)
mgus_haz_df <- int_info(mgus1_ped) |> 
  mutate(offset = intlen) |> 
  add_hazard(pam) 
```
]

.pull-right[
```{r, echo=FALSE, eval=TRUE, dependson=c("mgus-haz-df"), fig.width=5, fig.height=3, out.width="500px"}
ggplot(mgus_haz_df, aes(x = tend, y=hazard)) +
  geom_line() + ylim(c(0, 0.001)) + xlim(c(0,9000)) +
  geom_stepribbon(aes(ymin = ci_lower, ymax= ci_upper), alpha = .3) + xlab("time")
```

```{r, echo=TRUE, eval=FALSE}
ggplot(mgus_haz_df, aes(x = tend, y=hazard)) +
  geom_line() +
  geom_stepribbon(aes(ymin = ci_lower, 
                      ymax= ci_upper)
                  , alpha = .3) + 
  xlab("time") + 
  ylim(c(0, 0.001)) + xlim(c(0,9000))
```
]
