
---
class: inverse, middle, center

.font150[
**Part II: Intro to PEMs and PAMMs**
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900px>
</html>


```{r input-setup-pem, child="setup-pem.Rmd", eval = TRUE}
```


---
# Piecewise exponential model (PEM)

**Cox PH model:** $$h(t|\bfx_i) = h_0(t) \exp(\bfx_i^\top\bsbeta) = \exp(g_0(t) + \bfx_i^\top \bsbeta)$$
where the baseline hazard is written as $g_0(t) = \log h_0(t)$. <br>
We now want to estimate both $h_0(t)$ as well as the parameters
$\bsbeta$ via ML.

We decompose the time axis into $J$ intervals, using cut points $a_0, a_1, \ldots, a_J$:
$$(0=a_0,a_1],(a_1,a_2],\ldots,(a_{j-1},a_j],\ldots,(a_{J-1},a_J]$$
These interval cut points will often correspond to event and/or censoring times, but don't need to.

We assume piecewise constant hazards
$$h_0(t) = h_{0j} \quad \forall\, t \in I_j:=(a_{j-1},a_j]$$
with $\beta_{0j} = \log(h_{0j})$, this yields
$$h(t|\bfx_i) = h_{0j}\exp(\bfx_i^\top\bsbeta) = \exp(\beta_{0j} + \bfx_i^\top \bsbeta) = \exp(\eta_{ij})=: h_{ij} \quad \forall t \in I_j$$

---
# PEM Likelihoood
The full likelihood for right-censored data is then
$$\begin{align*}
L(\bsbeta,\bsbeta_0)
  &= \prod_{i=1}^n f(t_i|\bfx_i)^{\delta_i} S(t_i|\bfx_i)^{1-\delta_i} = \prod_{i=1}^n h(t_i|\bfx_i)^{\delta_i} \exp
    \left(
      -\int_0^{t_i} h(s|\bfx_i) ds
    \right).
\end{align*}$$
<!--where $\bsbeta_0 = (\beta_{01},\ldots,\beta_{0J})$. <br>-->
Since $h(s|\bfx_i)$ is piecewise constant, the integral in the (log-)likelihood is easy to compute.  <br>

We define pseudo-data

- event in interval $I_j$: $\delta_{ij} = \begin{cases} 1 & t_i \in I_j \mbox{ and } \ \delta_i = 1\\ 0 & \mbox{ else} \end{cases}$

- time at risk in interval $I_j$: $t_{ij} = \begin{cases} a_j - a_{j-1} & a_j < t_i\\ t_i - a_{j-1} & a_{j-1} < t_i \leq a_j\\ 0 & t_i \leq a_{j-1} \end{cases}$

- "offset": $o_{ij} = \log t_{ij} \quad (o_{ij} = -\infty \mbox{ for } t_{ij} = 0)$

---
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped.svg", width = "400px" align="middle" >
</div>
</div>


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped1.svg", width = "400px" align="middle" >
</div>
</div>



---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped2.svg", width = "400px" align="middle" >
</div>
</div>

- define: $\delta_{ij} = \begin{cases}1 & t_i \in I_j \text{ and } \delta_i = 1\\0 & \text{else}\end{cases}$

---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped3.svg", width = "400px" align="middle" >
</div>
</div>

- define: $\delta_{ij} = \begin{cases}1 & t_i \in I_j \text{ and } \delta_i = 1\\0 & \text{else}\end{cases}$,
$\quad t_{ij} = \begin{cases} a_j - a_{j-1} & a_j < t_i\\ t_i - a_{j-1} & a_{j-1} < t_i \leq a_j\end{cases}$



---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped4.svg", width = "400px" align="middle" >
</div>
</div>

- define: $\delta_{ij} = \begin{cases}1 & t_i \in I_j \text{ and } \delta_i = 1\\0 & \text{else}\end{cases}$,
$\quad t_{ij} = \begin{cases} a_j - a_{j-1} & a_j < t_i\\ t_i - a_{j-1} & a_{j-1} < t_i \leq a_j\end{cases}$,
$\quad t_j := a_j$


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped.svg", width = "400px" align="middle" >
</div>
</div>

- define: $\delta_{ij} = \begin{cases}1 & t_i \in I_j \text{ and } \delta_i = 1\\0 & \text{else}\end{cases}$,
$\quad t_{ij} = \begin{cases} a_j - a_{j-1} & a_j < t_i\\ t_i - a_{j-1} & a_{j-1} < t_i \leq a_j\end{cases}$,
$\quad t_j := a_j$

- also known as "start-stop" data format, as used for interval-censoring and time-varying covariates in PH models

---
# PEM Likelihoood (2)

Remember: $h(t|\bfx_i) = \exp(\beta_{0j} + \bfx_i^\top \bsbeta) =  \exp(\eta_{ij})$, so $h(t_i|\bfx_i)^{\delta_i} = \prod^J_{j=1} \exp \left( \delta_{ij} \eta_{ij}\right)= \prod^{J_i}_{j=1} \exp \left( \delta_{ij} \eta_{ij}\right).$

$J_i$ is the interval for which $t_i \in I_{J_i}=(a_{J_i-1},a_{J_i}]$, so
$$\int_0^{t_i} h(s|\bfx_i) ds
  = \sum^{J_i}_{j=1} t_{ij}\exp\left(\eta_{ij}\right) =  \sum^{J_i}_{j=1} \exp\left(o_{ij} + \eta_{ij}\right).$$
We can now rewrite the likelihood as:
$$L(\bsbeta,\bsbeta_0) = \prod_{i=1}^n \prod_{j=1}^{J_i}
  \exp \left( \delta_{ij} \eta_{ij} - \exp(o_{ij} + \eta_{ij}) \right)$$
with log-likelihood
$$l(\bsbeta,\bsbeta_0) = \log L(\bsbeta,\bsbeta_0) = \sum_{i=1}^n \sum_{j=1}^{J_i} \left(\delta_{ij} \eta_{ij} - \exp(o_{ij} + \eta_{ij}) \right)$$

---
# PEM Likelihood (3)

$$\text{Log-likelihood: }l(\bsbeta,\bsbeta_0) = \log L(\bsbeta,\bsbeta_0) = \sum_{i=1}^n \sum_{j=1}^{J_i} \left(\delta_{ij} \eta_{ij} - \exp(o_{ij} + \eta_{ij}) \right)$$

Now assume $\delta_{ij} \stackrel{iid}{\sim} Po(\mu_{ij})$, with $\mu_{ij} = h_{ij}t_{ij}$ and density $f(\delta_{ij}) = \tfrac{\mu_{ij}^{\delta_{ij}}}{\delta_{ij}!} \exp(-\mu_{ij})$

$$\begin{align*}
l_{Po}(\bsbeta,\bsbeta_0)
  &= \log\left(\prod_{i=1}^n\prod_{j=1}^{J_i}f(\delta_{ij})\right)=\sum_{i=1}^n\sum_{j=1}^{J_i}(\delta_{ij}\log(\mu_{ij}) - \mu_{ij})\\
  &= \sumn\sum_{j=1}^{J_i}(\delta_{ij}\log(h_{ij})+\delta_{ij}\log(t_{ij}) - h_{ij}t_{ij})\\
  & = \sumn\sum_{j=1}^{J_i}(\delta_{ij}\eta_{ij} - \exp(o_{ij} + \eta_{ij}) + \delta_{ij}o_{ij})
\end{align*}$$

$$\Rightarrow l_{Po}(\bsbeta, \bsbeta_0) \propto l(\bsbeta, \bsbeta_0)$$

$\ra$ We can fit a Poisson model to the pseudo-data to obtain ML estimates of $\bsbeta$ `r Citep(bib, c("holford_analysis_1980", "laird_covariance_1981", "friedman_piecewise_1982"))`

---
# PEM: Properties

**Trade-off**:
  - small $J$:
    + crude approximation of the baseline hazard
    + low computational cost
    + stable
  - large $J$:
    + more flexible approximation
    + high computational cost
    + unstable

In general
  - Number of baseline parameters to estimate equal to $J$
  - Number and placement of cut points $a_j, j=1,\ldots,J$ important for fit
  - If no ties in the data and cut points equal event times, $\hat{\bsbeta}_{PEM}=\hat{\bsbeta}_{Cox}$ `r Citep(bib, "whitehead_fitting_1980")`


---
# PEM: Illustration

```{r, est-pem, echo=FALSE, fig.width = 6, fig.height=3, dependson=c("sim-wb", "ex-pem-1")}
ped2 <- as_ped(Surv(time, status)~., cut = seq(0,10, by=.1), data=sim_df_weibull)
pem2 <- glm(ped_status ~ interval, data = ped2, family=poisson(), offset=offset)
pem_haz_df2 <- int_info(ped2) %>%
  mutate(
    hazard = predict(object=pem2, ., type="response"),
    survival = exp(-cumsum(hazard * intlen)))
p_pem_haz2 <- p_pem_haz %+% pem_haz_df2 + ylim(c(0, .35))
```

```{r, echo=FALSE, fig.width = 6, fig.height = 3, dependson=c("sim-wb", "ex-pem-1"), out.width = "800px"}
pem_haz1 + ylim(c(0, .35)) + ggtitle("PEM: small J") + p_pem_haz2 + ggtitle("PEM: large J")
```

--

**Solution**:
- use large $J$ and penalize differences between neighboring baseline hazards $\beta_{0j}, \beta_{0(j\pm1)}$ and/or

- reduce parameter count by *basis function representation* of $h_0(t)$.

---
# PEM: Penalized Likelihood

**Penalization:**

Impose a penalty on the step heights $$\operatorname{pen}(\bsbeta_0) = \sum_{j=2}^J (\beta_{0j} - \beta_{0j-1})^2,$$
this leads to the penalized estimator
$$(\widehat{\bsbeta}, \widehat{\bsbeta}_0) = \underset{\bsbeta,\bsbeta_0}{\mbox{argmax}} \ l_{pen}(\bsbeta,\bsbeta_0)$$
with
$$l_{pen}(\bsbeta,\bsbeta_0) = l(\bsbeta,\bsbeta_0) - \tau \ \operatorname{pen}(\bsbeta_0)$$ and
penalty/smoothing parameter $\tau$.

---
# PEM: Penalized Likelihood (2)

For
- $\tau \longrightarrow 0$ we obtain an unpenalized fit
- $\tau \longrightarrow \infty$ we obtain $g_0(t)=\text{const}$

$\ra$ It is therefore necessary to choose $\tau$ in a data-driven way.

$\ra$ As we simply optimise a (penalized) Poisson log-likelihood, we can use any optimization strategy / implementation for (penalized) GLMs or GAMMs:
**`pammtools`** provides lots of convenience functions to preprocess data and postprocess model outputs from e.g. **`mgcv`**.

PEM-representation also allows us to re-use *any* algorithm for time-to-event problems, as long as it can handle Poisson data with offsets (e.g. **`XGboost`** etc, see <a name=cite-bendergeneralMl2021></a>[Bender, Rügamer, Scheipl, and Bischl (2021)](#bib-bendergeneralMl2021)).

---
# PEM: Penalized Likelihood & Basis Representation

**Basis Representation:**

- Piecewise-constant log-baseline hazard rate can also be parameterized in terms of a spline basis, where $$g_0(t) = \sum_{\ell=1}^L \gamma_\ell B_\ell(t_j)$$
with $t_j = a_j\ \forall\, t \in I_j$, $L \ll J$, and typically a first order differences penalty on $\bsgamma$.

- *PEMs with general additive predictor:*
Since we're fitting a Poisson-Likelihood, this can *easily* be extended by including spatial, non-linear, random or time-varying effects, e.g.
$$h(t|\bfx_i,\bfz_i) = \exp(g_0(t) + f_1(z_{i1}) + \ldots + f_q(z_{iq}, t) +f_r(t)z_{ir} +  \bfx_i^\top\bsbeta).$$

$\ra$ In analogy to Piecewise Exponential Models (PEMs), estimated via GLMs, we refer to this model class as Piecewise-exponential Additive Mixed Model (PAMM, `r Citet(bib, "bender_generalized_2018")`), estimated via GAMMs


---
# PAM: Illustration
```{r, est-pam, echo=FALSE, fig.width = 6, fig.height=3, dependson=c("est-pem")}
pam <- mgcv::gam(ped_status ~ s(tend), data = ped2, family=poisson(), offset=offset)
pem_haz_df2 <- pem_haz_df2 |> add_hazard(pam, overwrite = TRUE)
p_pam_haz <- p_pem_haz %+% pem_haz_df2 +
  geom_stepribbon(aes(ymin = ci_lower, ymax= ci_upper), alpha = .3) + ylim(c(0, .35))
```

```{r, echo=FALSE, fig.width = 9, fig.height = 3, dependson=c("est-pam"), out.width = "800px"}
pem_haz1 + ylim(c(0, .35)) + ggtitle("PEM: small J") + p_pem_haz2 + ggtitle("PEM: large J")  + p_pam_haz + ggtitle("PAM: large J")
```

- A good strategy is to select cut points at observed event times $t_{(1)},\ldots,t_{(m)}$ (basis functions will be evaluated in regions with many events, i.e., regions with a lot of information about the process)
- For large $m$, a small subsample will often be sufficient
- Number of parameters $L$ does not depend on number of cut points $J$ (but computation time still affected by $J$)<br><br>
- PEMs and PAMMs in **R** facilitated by package **`pammtools`**
- Various articles (vignettes) on different topics available at [adibender.github.io/pammtools](https://adibender.github.io/pammtools/articles/)

---
# **`pammtools`**: Important utility functions

*Pre-Processing*:

- `as_ped(data, formula, cut, ...)` creates PED-formatted data suitable for PAMMs.
Also merges additional datasets for time-dependent covariates according to the specification in `formula` and handles left-truncation.

*Post-Processing*:

- `make_newdata(x, <optional covariate specifications>)` creates PED-formatted data suitable for evaluating PAMM estimates for all intervals using `add_???`-functions.
  Covariates are set to their sample means or modes by default.
- `add_???(newdata, object)` computes PAMM point estimates (and CIs) from a model `object` on `newdata` and appends them to `newdata`.  Resulting `data.frame` useful for custom visualizations etc.
   Flavors: `add_surv_prob`, `add_hazard`, `add_cumu_hazard`, `add_cif`, `add_term`
- `tidy_fixed()`, `tidy_re()`, `tidy_smooth()` extract estimated coefficients / smooth effects and CIs as clean `data.frames`
- `gg_fixed()`, `gg_smooth()`, `gg_slice()`, ... etc for direct `ggplot2` viz of estimates
