

---
class: inverse, middle, center

.font150[
**2. How can `pammtools` help me analyze survival data?** <br>
A quick overview to get from survival data to interpretable results
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900px>
</html>

---
# `pammtools` exploits the benefits of piecewise exponential additive mixed models (PAMMs)

**Why exponential additive mixed model?**
- Same "hazard structure" as in Cox PH Models $\implies$ exponential
- BUT can make use of GAMM advantages, e.g. smooth effects ($f$) dependent on $x$ and $t$, mixed effects $b$ for clusters $\ell$
$$\log(h_{k,e}(t|\mathbf{x}_{i, k, e}, \ell_i)) = \beta_{0,k,e} + f_{0,k,e}(t) + \sum_{p=1}^P f_p(x_{i,k,e,p}, t) + b_{\ell_i}$$
where the baseline hazard is written as $\beta_{0,k,e} + f_{0,k,e}(t)$. <br>
We now want to estimate both $h_0(t)$ as well as the parameters
$\bsbeta$ via ML.

**Why piecewise?**
- Decompose the time axis into intervals
- Assume piecewise constant hazards in these intervals

**What data do we need for that?**


---
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped.svg", width = "400px" align="middle" >
</div>
</div>


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped1.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped2.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval

---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped3.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval
- Calculate differences of event times $t_i$ and interval endpoints $a_j$


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped4.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval
- Calculate differences of event times $t_i$ and interval endpoints $a_j$
- Define the "new" time points to the interval endpoints


---
count: false
# Piecewise Exponential Data (PED)

<div class="row">
<div class = "column", align = "center">
Data in "standard" time-to-event format <br>
</div>
<div class = "column", align = "center">
Data in PED format <br>
</div>
</div>

<div class = "row" align = "middle">
<div class = "column", align = "middle">
.middle[
<img src="figures/tab-standard.svg", width = "300px" align="middle"><br>
$\ra$ transform to PED using $a_0=0, a_1 = 1, a_2=1.5, a_3=3$
]
</div>

<div class = "column" align ="middle">
<img src="figures/tab-ped.svg", width = "400px" align="middle" >
</div>
</div>

- Include information on interval endpoints $a_j$
- Set the status to $1$ if an event happens within the time interval
- Calculate differences of event times $t_i$ and interval endpoints $a_j$
- Define the "new" time points to the interval endpoints

Note: Often interval endpoints are event times.

---
# Use `pammtools` to get from survival data to illustrations

.pull-left[
Example (Baseline): Survival data of tumor patients after operation `tumor` from the `pammtools` package
- 776 observations
- `status`: 0/1 for censoring or death
- `days`: days from operation to event

```{r, echo=TRUE, eval=FALSE}
data(tumor, package="survival")
tumor_ped <- tumor |> as_ped(Surv(days, status)~.)
pam <- mgcv::gam(
  formula = ped_status ~ s(tend), data=tumor_ped,
  family=poisson(),offset = offset)
# shorter
pam <- pamm(ped_status ~ s(tend), data = tumor_ped)
haz_df <- tumor_ped |>
  make_newdata(tend = unique(tend)) |>
  add_hazard(pam)
```
]

.pull-right[
```{r, echo=FALSE, eval=TRUE, dependson=c("mgus-haz-df"), fig.width=5, fig.height=3, out.width="500px"}
ggplot(haz_df, aes(x=tend, y=hazard)) +
  geom_line() +
  geom_stepribbon(aes(ymin = ci_lower, ymax= ci_upper), alpha = .3) + xlab("time")
```

```{r, echo=TRUE, eval=FALSE}
ggplot(haz_df, aes(x=tend, y=hazard)) +
  geom_line() +
  geom_stepribbon(aes(ymin = ci_lower, ymax= ci_upper), alpha = .3) +
  xlab("time")
```
]
