
---
class: inverse, middle, center

.center[
.font120[
**Non-proportional Hazards**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>



---
# Modelling Time-Varying Effects (TVEs): Setting & PAMMs applicability

**Setting**: 
The effect of a covariate $x$ can change as time progresses, e.g. effect of patients' status at admission to the hospital vanishes over time,

**Main idea**:
Define *artificial time-dependent covariates* and fit the model with these time-dependent covariates to represent time-varying effects (TVEs).

$$\beta(t)=\sum^L_{\ell=1} \gamma_\ell B_\ell(t)$$

- **In PAMMs**, very easy to include since we already include time $t$ as a covariate for the baseline hazard
- Time-varying linear effects $x\beta(t)$ simply specified as spline interaction effect `s(<TIME>, by = <X>)`
- Non-linear time-varying effects $f(x, t)$ specified as tensor product splines `te(<TIME>, <X>)`

Technical point:
- PAMMs assume piece-wise constant hazard rates, so $\beta(t)$ actually modelled as a *step function* that only changes at cutpoints $a_j$:
$\beta(t) \equiv \sum^L_{\ell=1} \gamma_\ell B_\ell(a_{j})\; \forall\, t \in (a_{j-1}, a_j]$
-  for data created with `as_ped`, `tend` is $a_j$:  use `s(tend, by = <X>)`

---
# TVEs - Example: Veterans' Administration Lung Cancer study

**Recap:** Randomised trial of survival times under two treatment regimens for lung cancer
- endpoint: `status` (censoring = 0)
- variables of interest:  `karno` Karnovsky performance score (consider as TVE), `trt` treatment (standard = 1, test = 0), `prior` prior therapy (no = 0, yes = 10)
```{r vet-load, echo = TRUE}
veteran <- survival::veteran  |> filter(time < 400)  |> #reduce data for ex
  mutate(trt   = 1L * (trt == 2),
         prior = 1L * (prior == 10))
rbind(head(survival::veteran, n = 2), tail(survival::veteran, n = 2))[, c(1:5, 8)] |> knitr::kable()
```

---
# TVEs - Example: Fixed shape vs. Flexible time-varying effects (1)

Instead of linear effect, literature suggests time-varying effect of `karno` with
$f(x_{\text{karno}},t) = \beta_{\text{karno}} x_{\text{karno}} + \beta_{\text{karno},t}x_{\text{karno}}\log(t+20)$

PAM specification:
```{r vet-pam, echo =TRUE}
# data transformation
veteran_ped <- veteran |> as_ped(Surv(time, status)~., id = "id") |>
    mutate(logt20 = log(tend + 20))
# fit model
vet_pam_logt <- pamm(ped_status ~ s(tend) + trt + prior + karno + karno:logt20,
  data = veteran_ped)
```


---
# TVEs - Example: Fixed shape vs. Flexible time-varying effects (2)

Alternatively, specify the effect of `karno` as
$f(x_{\text{karno}},t) = f(t)\cdot x_{\text{karno}}$, where $f(t)$
is estimated from the data:
```{r vet-pam-tve, echo =TRUE}
# no need for separate time-constant effect for karno here:
vet_pam_tve <- pamm(
  ped_status ~ s(tend) + trt + prior + s(tend, by = karno),
  data = veteran_ped)
AIC(vet_pam_logt, vet_pam_tve)
```
$\implies$ assumed shape "good enough" in this case, Tensor spline has minimal AIC

---
# Example: Flexible time-varying effects (2)
```{r gg_tv_karno, echo =TRUE, fig.height = 3, fig.width = 5, out.height= "300px"}
term_df <- veteran_ped |> ped_info() |> add_term(vet_pam_tve, term = "karno") |>
  mutate_at(c("fit", "ci_lower", "ci_upper"), funs(. / .data$karno)) |>
  mutate(pam.fit = coef(vet_pam_logt)["karno"] + coef(vet_pam_logt)["karno:logt20"] * log(tend + 20))
ggplot(term_df, aes(x = tend, y = fit)) + geom_step(aes(col = "PAM spline")) +
    geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
    geom_step(aes(y = pam.fit, col = "PAM log(t)")) +
    scale_color_manual(name = "Model: ", values = c(Set1[1:2], "black")) +
    xlab("t") + ylab(expression(hat(beta)(t)))
```

---
# Stratification: Time-varying Effects of Factors

$$h(t | x_i)= h_0(t)\exp(x_i\beta(t))$$

If $x_i$ is a dummy variable for a level of a categorical variable,
this model is approximately equivalent to a *stratified* PH model with a
different "baseline" hazard for the category encoded by $x_i$.

- PAMM formula specification `strata + s(tend, by = strata)` with a factor variable `strata` yields different (baseline) hazard rates for each level of `strata`
- See `?mgcv::gam.models` and `?mgcv::factor.smoooth` for alternative penalization and centering schemes if `strata` has many levels
- Also `r Citet(bib, "pedersen_hierarchical_2019")` for overview of hierarchical GAMs

---
# TVEs - Example: Stratification via factorial TVE

```{r vet-strat-cox, echo =TRUE, fig.height = 3, fig.width = 5, out.height=".4\textheight", warning = TRUE}
vet_cox_strata <- coxph(Surv(time, status) ~ strata(celltype),
                        data = veteran)
vet_pam_strata <- pamm(ped_status ~ s(tend, by = celltype) + celltype,
                       data = veteran_ped)

# extract fitted cumulative hazards:
pam_fit <- veteran_ped |>
  ped_info() |> slice(rep(1:n(), each = 4)) |>
  mutate(celltype = rep(levels(celltype), times = n()/4)) |>
  group_by(celltype) |> arrange(tend)  |>  #!!
  add_cumu_hazard(vet_pam_strata)
cox_fit <- vet_cox_strata |> basehaz()
```

---
# TVEs - Example: Stratification via factorial TVE (2)

```{r gg_vet_strata, echo =TRUE, fig.height = 5, fig.width = 7, out.width  = "600px",}
ggplot(pam_fit, aes(x = tend, y = cumu_hazard)) + geom_step(aes(col = celltype)) +
    geom_stepribbon(aes(ymin = cumu_lower, ymax = cumu_upper, fill = celltype), alpha = 0.1) +
    geom_line(data = cox_fit, aes(x = time, y = hazard, col = strata), linetype = 4, size = 1) +
    ylab(expression(hat(H)(t))) + xlab("t") + coord_cartesian(ylim = c(0, 10))
```

