
---
class: inverse, middle, center

.center[
.font120[
**Time-dependent Covariates**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---
# Time-Dependent Covariates (TDCs)

If a covariate $x$ changes over time $t$, the hazard rate becomes
$$h(t | x_i(t))=h_0(t)\exp(x_i(t)\beta).$$ 

Note that the resulting model **does not have proportional hazards**, since
$$\frac{\lambda(t, x_1(t))}{\lambda(t, x_2(t))}=\exp((x_1(t)-x_2(t))'\beta)$$
depends on $t$.

---
# Time-Dependent Covariates: Data Format

PAMMs assume piecewise-constant hazard rate  
$\implies$ need to assume piecewise constant $x(t) \equiv x\left(a_j\right)$ for $a_{j-1} < t \leq a_{j}$  
$\implies$ include all times at which $x_i$ changes for any $i$ as cut points

- `pammtools::as_ped`: use formula special term `concurrent` to denote time-varying covariates

- Computational effort for Cox PH models becomes a *lot* larger in start-stop format, remains unchanged for PAMMs.
- "Start-stop" data format is also used for interval-censored and left-truncated data, c.f. `survival::survSplit`, `survival::tmerge`


---
# Example: Primary Biliary Cirrhosis

.pull-left[ 
`survival::pbc`:
```{r pbc-1, echo = FALSE}
data("pbc", package = "survival")
head(survival::pbc)[, c(1:5, 11, 12)] |> knitr::kable()
```
]
.pull-right[ 
`survival::pbcseq`:  
```{r pbc-2, echo = FALSE}
head(survival::pbcseq)[, c(1, 4:5, 7, 12, 18)] |> knitr::kable()
```
]

---
# Example: Primary Biliary Cirrhosis (2)

Data transformation using `as_ped` function with formula special `concurrent`:

```{r pbc-3, echo = TRUE}
pbc_clean <- pbc  |>  
  mutate(status = 1L * (status == 2)) |>
  filter(id <= 312) |>  select(id:sex) 
pbc_seq <- pbcseq |>  
  mutate(log_bili = log(bili), log_protime = log(protime)) |>  
  select(id, day, log_bili, log_protime)  
pbc_ped <- as_ped(
  data = list(pbc_clean, pbc_seq), #!!
  formula = Surv(time, status) ~ . + concurrent(log_bili, log_protime, tz_var = "day"), #!!
  id = "id")
glimpse(pbc_ped)
```

---
# Example: Primary Biliary Cirrhosis (4)

.pull-left[ 
`pbcseq`:  
```{r pbc-2b, echo = FALSE}
head(pbc_seq) |> knitr::kable()
```
]

.pull-right[ 
`pbc_ped[49:54, ]` via `pammtools::as_ped`:  
```{r pbc-4, echo = FALSE}
pbc_ped[49:54, -c(4, 5)] |> knitr::kable(row.names = FALSE)
```
]

---
# Instantaneous and delayed effects of TDCs

Default is *concurrent* or instantaneous effect:  

- covariate value $x_i(t)$ observed at $t = a_j$ instantaneously affects the
hazard starting from interval $(a_j, a_{j+1}]$ until the next change in $x$ or end of follow-up. 

*Lagged* effects:

- use  `lag` argument in `concurrent` to specify time-delayed effect of $x_i(t)$
- covariate value $x_i(t)$ observed at $t = a_j$ affects the
hazard starting from later interval $(a_j + \texttt{lag}, \ldots]$ until the next (lagged) change in $x$ or end of follow-up.

Much more generally, **pammtools** can also be used to fit delayed and time-limited, cumulative and time-varying effects of TDCs, so called
exposure-lag response associations (ELRA, `r Citet(bib, "bender_penalized_2019")`)

---
# Extended Cox analysis of the `pbc` data
```{r pbc-cox1, echo = TRUE}
# code adapted from vignette("timedep", package="survival")
temp <- subset(pbc_clean, select = c(id:sex)) # baseline
pbc2 <- survival::tmerge(temp, temp, id = id, death = event(time, status)) #set range
pbc2 <- survival::tmerge(pbc2, pbc_seq, id = id, log_bili = tdc(day, log_bili),
  log_protime = tdc(day, log_protime))
```
.pull-left[ 
Use baseline measurements only:
```{r pbc-cox2, echo = TRUE}
fit1 <- coxph(Surv(time, status == 2) ~ 
                log(bili) + log(protime), 
              data = subset(pbc, id <= 312))
cbind(coef(fit1), confint(fit1)) |> exp() |> kable()
```
]
.pull-right[ 
Use all measurements:
```{r pbc-cox3, echo = TRUE}
fit2 <- coxph(Surv(tstart, tstop, death == 1) ~ 
                log_bili + log_protime, 
              data = pbc2)
cbind(coef(fit2), confint(fit2)) |> exp() |> kable()
```
]
.center[
$\implies$ substantial differences
]
---
# PAM analysis of the `pbc` data
Fit `PEM`-model using `glm`:

```{r pbc-pam, echo  =TRUE, eval = TRUE, cache= TRUE}
pbc_pam <- mgcv::gam(ped_status ~ s(tend) + log_bili + log_protime, data = pbc_ped,
  family = poisson(), offset = offset)
summary(pbc_pam)

coef(pbc_pam)[2:3]
coef(fit2)
```

---
# PAM analysis of the `pbc` data (2)
Coefficient estimates are similar, especially for the
effect of `bili`:
```{r pbc-fig-prep,  fig.keep="none"}
## Effect of bilirubin
# note that we use the reference argument to calculate
# the relative risk change (x - \bar{x})'\beta for comparison with predict.coxph
# (see also Details section in ?predict.coxph)
reference = sample_info(pbc_ped)
bili_df <- pbc_ped %>% ungroup() %>%
  make_newdata(log_bili = seq_range(log_bili, n = 100)) %>%
  add_term(pbc_pam, term = "log_bili", reference = reference) %>%
  mutate(cox = predict(fit2, ., type = "term")[, "log_bili"])

## Effect of protime
protime_df <- pbc_ped %>% ungroup() %>%
  make_newdata(log_protime = seq_range(log_protime, n=100)) %>%
  add_term(pbc_pam, term = "log_protime", reference = reference) %>%
  mutate(cox = predict(fit2, ., type = "term")[, "log_protime"])

# visualization
# remember that bili and protime are log transformed
p_term <- ggplot(data = NULL, aes(y = fit)) + geom_line(aes(col = "PAM")) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
  geom_line(aes(y = cox, col = "Cox")) +
  scale_colour_manual(name = "Method", values = c("#E41A1C", "#000000"))
gridExtra::grid.arrange(
  p_term %+% bili_df + aes(x = exp(log_bili)) + xlab("Bilirubin") + ylab("ln(HR)"),
  p_term %+% protime_df + aes(x = exp(log_protime))+ xlab("Protime") + ylab("ln(HR)") + 
    labs(caption = "vertical axis shows ln(HR) compared to sample average"),
  nrow = 1L)
```

```{r pbc-fig, echo = FALSE, fig.width = 6, fig.height = 3}
gridExtra::grid.arrange(
  p_term %+% bili_df + aes(x = exp(bili)),
  p_term %+% protime_df + aes(x = exp(protime)),
  nrow = 1L)
```
