
---
count: false
class: inverse, middle, center

.center[
.font120[
**Non-proportional Hazards**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---
count: false
# Time-Dependent Covariates (TDCs): Setting & PAMMs applicability

Setting: covariate $x$ changes over time $t$  <br>
$\implies$ the hazard rate becomes time dependent  <br>
$\implies$ the hazard ratio becomes time dependent, i.e. the resulting model **does not have proportional hazards**

PAMMs assume piecewise-constant hazard rate <br>
$\implies$ need to assume piecewise constant $x(t) \equiv x\left(a_j\right)$ for $t \in (a_{j-1}, a_{j}]$ <br>
$\implies$ need to include all times at which $x_i$ changes for any $i$ as cut points

- `pammtools::as_ped`: use formula special term `concurrent` to denote time-varying covariates and include all necessary cut points.<br><br>
- Computational effort for Cox PH models becomes a *lot* larger in start-stop format, affects PAMMs less.
- "Start-stop" data format is also used for left-truncated data and time-dependent covariates in the Cox model, c.f. `survival::survSplit`, `survival::tmerge`


---
count: false
# TDCs - Example: Primary Biliary Cirrhosis

`pbc`: Mayo Clinic Trial between 1974 and 1984, `pbcseq`: follow-up laboratory data.

- 424 PBC patients
- endpoints: `status` 0/1/2 for censored, transplant, death 
- death (`status` = 2): event of interest (later: example on competing risks)
- variables of interest:  `bili` bilirunbin in blood serum(mg/dl) and `protime` standardised blood clotting time

.pull-left[
`survival::pbc`:
```{r pbc-1, echo = FALSE}
data("pbc", package = "survival")
head(survival::pbc, n=4)[, c(1:5, 11, 12)] |> knitr::kable()
```
]
.pull-right[
`survival::pbcseq`:
```{r pbc-2, echo = FALSE}
head(survival::pbcseq, n = 4)[, c(1, 4:5, 7, 12, 18)] |> knitr::kable()
```
]

---
count: false
# TDCs - Example: Primary Biliary Cirrhosis (2)

Data transformation using `as_ped` function with formula special `concurrent`:

```{r pbc-3, echo = TRUE}
pbc_clean <- pbc  |>  mutate(status = 1L * (status == 2)) |>
  filter(id <= 312) |>  select(id:sex)
pbc_seq <- pbcseq |>  mutate(log_bili = log(bili), log_protime = log(protime)) |>
  select(id, day, log_bili, log_protime)
pbc_ped <- as_ped(
  data = list(pbc_clean, pbc_seq), id = "id",  #!!
  formula = Surv(time, status) ~ . + concurrent(log_bili, log_protime, tz_var = "day")) #!!
```

This creates splits at
+ each event time
+ each time-point where the value of a TDC is updated




---
count: false
# TDCs - PAM analysis of the `pbc` data
```{r pbc-pam, echo  =TRUE, eval = TRUE, cache= TRUE, output.lines = c(8:18)}
pbc_pam <- pamm(ped_status ~ s(tend) + log_bili + log_protime, data = pbc_ped)
summary(pbc_pam)
```

---
count: false
# TDCs - PAM analysis of the `pbc` data (2)

```{r pbc-cox1, echo = FALSE}
# code adapted from vignette("timedep", package="survival")
temp <- subset(pbc_clean, select = c(id:sex)) # baseline
pbc2 <- survival::tmerge(temp, temp, id = id, death = event(time, status)) #set range
pbc2 <- survival::tmerge(pbc2, pbc_seq, id = id, log_bili = tdc(day, log_bili),
                         log_protime = tdc(day, log_protime))
```

```{r pbc-cox2, echo = FALSE}
pbc_cox_base <- coxph(Surv(time, status == 2) ~
                log(bili) + log(protime),
              data = subset(pbc, id <= 312))
```

```{r pbc-cox3, echo = FALSE}
pbc_cox_tdc <- coxph(Surv(tstart, tstop, death == 1) ~
                log_bili + log_protime,
              data = pbc2)
```

```{r pbc-pam2, echo  = TRUE, eval = FALSE, cache= TRUE}
pammtools::pamm(ped_status ~ s(tend) + log_bili + log_protime,
                data = pbc_ped)
#equivalent to:
mgcv::gam(ped_status ~ s(tend) + log_bili + log_protime,
          data = pbc_ped,
          family = poisson(), offset = offset)
```

Similar results for Cox and PAMM models:
```{r pbc-pam3, echo  = TRUE}
coef(pbc_pam)[2:3]
coef(pbc_cox_tdc)
```

---
count: false
# TDCs - PAM analysis of the `pbc` data (3)
Coefficient estimates are similar, especially for the
effect of `bili`:
```{r pbc-fig,  fig.width = 9, fig.height = 4, out.width = "\\textwidth"}
## Effect of bilirubin
# note that we use the reference argument to calculate
# the relative risk change (x - \bar{x})'\beta for comparison with predict.coxph
# (see also Details section in ?predict.coxph)
reference = sample_info(pbc_ped)
bili_df <- pbc_ped |> ungroup() |>
  make_newdata(log_bili = seq_range(log_bili, n = 100)) |>
  add_term(pbc_pam, term = "log_bili", reference = reference)
bili_df$cox <- predict(pbc_cox_tdc, bili_df, type = "term")[, "log_bili"]

## Effect of protime
protime_df <- pbc_ped |> ungroup() |>
  make_newdata(log_protime = seq_range(log_protime, n=100)) |>
  add_term(pbc_pam, term = "log_protime", reference = reference)
protime_df$cox <- predict(pbc_cox_tdc, protime_df, type = "term")[, "log_protime"]

# visualization
# remember that bili and protime are log transformed
p_term <- ggplot(data = NULL, aes(y = fit)) + geom_line(aes(col = "PAM")) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
  geom_line(aes(y = cox, col = "Cox")) +
  scale_colour_manual(name = "Method", values = c("#E41A1C", "#000000"))

(p_term %+% bili_df + aes(x = exp(log_bili)) + xlab("Bilirubin") + ylab("ln(HR)") + theme(legend.position = "none")) +
(p_term %+% protime_df + aes(x = exp(log_protime))+ xlab("Protime") + ylab("ln(HR)")) +
patchwork::plot_annotation(caption = "vertical axis shows ln(HR) w.r.t. sample average")
```