
---
class: inverse, middle, center

.center[
.font120[
**Time-varying effects**
]
]

<html>
  <div style=---float:left></div>
  <hr color='#005500' size=1px width=900pxpx>
</html>


---

# Time-Varying Effects

The effect of a covariate $x$ can change as time progresses

Examples:

- effect of patients' status at admission to the hospital vanishes over time,
- whether a tool is made with cheap or high-quality materials begins to affect its risk of failure only after it has been used for a certain amount of time.

**Model for one covariate**
$$h(t | x_i)= h_0(t)\exp(x_i\beta(t))$$ 

Note that the proportional hazard assumption is no longer fulfilled since
$$\frac{h(t | x)}{h(t | x + 1)} = \exp(\beta(t)):$$ 
$\implies$ The hazard ratio becomes a function of time.

---
# Modelling Time-Varying Effects: Idea

Main idea:  
Define *artificial time-dependent covariates* and fit the model with these time-dependent covariates to represent time-varying effects (TVEs).

Specifically:  
$$\beta(t)=\sum^L_{\ell=1} \gamma_\ell B_\ell(t)\ ,$$ with basis functions
depending on t:
$$h(t |x_i)= h_0(t)\exp\bigg(x_i \sum_\ell \gamma_\ell B_\ell(t)\bigg)=h_0(t)\exp\bigg(\sum_\ell \gamma_\ell (x_i B_\ell(t))\bigg).$$
$\implies$ coefficients $\gamma_\ell$ for time-dependent covariates $x B_\ell(t)$ can be estimated as usual

---
# Modelling Time-Varying Effects (1)

Examples:

- $L=1, B_1(t) \equiv 1$: time-constant effect
- $L=1, B_1(t) = \log(1 + t)$: logarithmic increase/decrease over time

Much more flexible & fairly assumption-free:  
- Use $L \gg 1$ 
- Use spline basis functions for $B_\ell(t)$
- Penalize $\gamma_\ell$

---
# Modelling Time-Varying Effects (2)

$$\beta(t)=\sum^L_{\ell=1} \gamma_\ell B_\ell(t)$$

- **Very** easy to include in PAMMs since we already include time $t$ as a covariate for the baseline hazard
- Time-varying linear effects $x\beta(t)$ simply specified as spline interaction effect `s(<TIME>, by = <X>)` 
- Non-linear time-varying effects $f(x, t)$ specified as tensor product splines `te(<TIME>, <X>)`

Technical point:  
- PAMMs assume piece-wise constant hazard rates, so $\beta(t)$ actually modelled as a *step function* that only changes at cutpoints $a_j$:  
$\beta(t) \equiv \sum^L_{\ell=1} \gamma_\ell B_\ell(a_{j})\; \forall\, t \in (a_{j-1}, a_j]$
-  for data created with `as_ped`, `tend` is $a_j$:  use `s(tend, by = <X>)`

---
# Example: Veterans' Administration Lung Cancer study 

Randomised trial of survival times under two treatment regimens for lung cancer:
```{r vet-load, echo = TRUE}
veteran <- survival::veteran  |> filter(time < 400)  |> #reduce data for ex
  mutate(trt   = 1L * (trt == 2),
         prior = 1L * (prior == 10))
str(veteran)
```
We consider TVEs of Karnofsky performance scores at randomisation (`karno`).

---
# Example: Fixed shape for time-variation (1)

Literature suggests time-varying effect of `karno` with  
$f(x_{\text{karno}},t) = \beta_{\text{karno}} x_{\text{karno}} + \beta_{\text{karno},t}x_{\text{karno}}\log(t+20)$

```{r vet-cox, echo =TRUE}
vet_cox_logt <- coxph(formula = Surv(time, status) ~ trt + prior + karno + tt(karno),
                     tt = function(x, t, ...) x * log(t + 20), #!!
                     data = veteran)
coef(vet_cox_logt)
```
```{r, echo=FALSE}
ttcoef <- round(coef(vet_cox_logt), 3)[3:4]
```
$\implies$ Estimated shape of time variation:  
$$\hat\beta_{\text{karno}}+\hat\beta_{\text{karno},t}\cdot\log(t+20) = `r ttcoef[1]` + `r ttcoef[2]`\cdot\log(t+20)$$

---
# Example: Fixed shape for time-variation (2)

.pull-left[
Equivalent PAM specification:
```{r vet-pam, echo =TRUE}
# data transformation
veteran_ped <- veteran |> 
    as_ped(Surv(time, status)~., id = "id") |>
    mutate(logt20 = log(tend + 20))
# fit model
vet_pam_logt <- pamm(
  ped_status ~ s(tend) + trt + prior + 
                 karno + karno:logt20, #!!
  data = veteran_ped
)
```
]
.pull-right[
Compare fits:
```{r vet-pam-coef, echo = TRUE}
cbind(
  pam = coef(vet_pam_logt)[2:5],
  cox = coef(vet_cox_logt)) |> kable()
```
]

---
# Example: Flexible time-varying effects

Specify the effect of `karno` as
$f(x_{\text{karno}},t) = f(t)\cdot x_{\text{karno}}$, where $f(t)$
is estimated from the data:
```{r vet-pam-tve, echo =TRUE}
# no need for separate time-constant effect for karno here:
vet_pam_tve <- pamm(
  ped_status ~ s(tend) + trt + prior + s(tend, by = karno),
  data = veteran_ped)
AIC(vet_pam_logt, 
    vet_pam_tve)
```
$\implies$ assumed shape "good enough" in this case

---
# Example: Flexible time-varying effects (2)
```{r gg_tv_karno, echo =TRUE, fig.height = 3, fig.width = 5, out.height= "300px"}
term_df <- veteran_ped |> ped_info() |> add_term(vet_pam_tve, term = "karno") |>
  mutate_at(c("fit", "ci_lower", "ci_upper"), funs(. / .data$karno)) |>
  mutate(cox.fit = coef(vet_cox_logt)["karno"] + coef(vet_cox_logt)["tt(karno)"] * log(tend + 20),
    pam.fit = coef(vet_pam_logt)["karno"] + coef(vet_pam_logt)["karno:logt20"] * log(tend + 20))
ggplot(term_df, aes(x = tend, y = fit)) + geom_step(aes(col = "PAM spline")) +
    geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
    geom_line(aes(y = cox.fit, col = "Cox log(t)")) + geom_step(aes(y = pam.fit, col = "PAM log(t)")) +
    scale_color_manual(name = "Model: ", values = c(Set1[1:2], "black")) +
    xlab("t") + ylab(expression(hat(beta)(t)))
```


---
# Stratification: Time-varying Effects of Factors

$$h(t | x_i)= h_0(t)\exp(x_i\beta(t))$$

If $x_i$ is a dummy variable for a level of a categorical variable, 
this model is approximately equivalent to a *stratified* PH model with a
different "baseline" hazard for the category encoded by $x_i$. 

- PAMM formula specification `strata + s(tend, by = strata)` with a factor variable `strata` yields different (baseline) hazard rates for each level of `strata`
- See `?mgcv::gam.models` and `?mgcv::factor.smoooth` for alternative penalization and centering schemes if `strata` has many levels.

---
# Example: Stratification via factorial TVE

```{r vet-strat-cox, echo =TRUE, fig.height = 3, fig.width = 5, out.height=".4\textheight", warning = TRUE}
vet_cox_strata <- coxph(Surv(time, status) ~ strata(celltype),
                        data = veteran)
vet_pam_strata <- pamm(ped_status ~ s(tend, by = celltype) + celltype,
                       data = veteran_ped)

# extract fitted cumulative hazards:
pam_fit <- veteran_ped |> 
  ped_info() |> slice(rep(1:n(), each = 4)) |> 
  mutate(celltype = rep(levels(celltype), times = n()/4)) |> 
  group_by(celltype) |> arrange(tend)  |>  #!!
  add_cumu_hazard(vet_pam_strata) 
cox_fit <- vet_cox_strata |> basehaz() 
```

---
# Example: Flexible time-varying effects (2)

```{r gg_vet_strata, echo =TRUE, fig.height = 5, fig.width = 7, out.width  = "600px",}
ggplot(pam_fit, aes(x = tend, y = cumu_hazard)) + geom_step(aes(col = celltype)) +  
    geom_stepribbon(aes(ymin = cumu_lower, ymax = cumu_upper, fill = celltype), alpha = 0.1) +
    geom_line(data = cox_fit, aes(x = time, y = hazard, col = strata), linetype = 4, size = 1) + 
    ylab(expression(hat(H)(t))) + xlab("t") + coord_cartesian(ylim = c(0, 10)) 
```

---
# Testing for Proportional Hazards / Time-Varying Effects

We can test the proportional hazards assumption with regard to specific covariates by checking for statistically significant time-varying effects:

$$h(t|x)=h_0(t) \exp(\beta_1 x+ \beta_2(t) x)$$ with $\beta_2(t) = \sum_\ell B_\ell(t)\gamma_\ell$

Check for $\beta_2(t) \equiv 0$ with $H_0: \gamma_\ell = 0 \,\forall\, \ell$

---
# Example: Testing for Proportional Hazards (1)

Simple z-Test for time-varying effect `karno:logt20`:
```{r, echo = TRUE, output.lines = c(1:14)}
summary(vet_pam_logt) 
```

---
# Example: Testing for Proportional Hazards (2)
Likelihod ratio test for smooth time-varying effect vs time-constant effect:
```{r, echo = TRUE}
vet_pam_const <- update(vet_pam_tve, .~. - s(tend, by = karno) + karno)
anova(vet_pam_tve, vet_pam_const, test = "Chisq") 
```

---
# Example: Testing for Proportional Hazards (3)
Careful:  
p-values of smooth terms in `summary.gam` are for $H_0:  f(x, t) \equiv 0$, not $H_0:  f(x, t) \equiv x \beta_1$:
```{r, echo = TRUE, output.lines = c(1:6, 15:20)}
summary(vet_pam_tve) 
```
