---
title: "IWSM 2023 - short course"
subtitle: "Exercise II: Competing Risks"
author: "Andreas Bender, Fabian Scheipl"
---

```{r, echo = FALSE}
options(width = 120)
library(knitr)
opts_chunk$set(
  fig.align = "center",
  warning   = FALSE
)
```



```{r, echo = TRUE, message = FALSE}
library(pammtools)
library(mgcv)
library(ggplot2)
theme_set(theme_bw())
library(survival)
solution <- TRUE
```


The package **`etm`** contains the `abortion` data set (Meister, R. and Schaefer, C. (2008). *Statistical methods for estimating the probability of spontaneous abortion in observational studies – Analyzing pregnancies exposed to coumarin derivatives*. Reproductive Toxicology, 26, 31–35).

It was collected in a prospective study. 1186 pregnant women were included in order to investigate spontaneous and induced abortions. 173 of them were exposed to *coumarin* derivatives (blood thinner), that are suspected to disturb the development of embryos or fetuses.

The research question of interest was whether and how exposure to *coumarin* affects the pregnancy outcome, which could be one of

- spontaneous abortion
- induced abortion
- live birth


1. Load and investigate the data

```{r, echo = solution, eval = solution}
data("abortion", package = "etm")
head(abortion)
table(abortion$cause)
prop.table(xtabs( ~ cause + group, data = abortion), margin = 2)
```

2. Consider the (implicit) criterion for women to be included into the study.
What does the analysis have to account for?


3. Estimate a PAM and compute the Cumulative incidence function (CIF) for both groups of women, i.e. the ones exposed to coumarin and the ones without exposure (specified by the `group` variable).


```{r, echo = solution, eval = solution}
abortion_ped <- abortion %>%
  mutate(
    cause = factor(
      cause,
      levels = 1:3,
      labels = c("induced", "life birth", "spontaneous")),
    group = factor(
      group,
      levels = 0:1,
      labels = c("control", "coumarin exposure"))
  ) %>%
  as_ped(Surv(entry, exit, cause) ~ ., id = "id", combine = TRUE)

# model with:
#   - cause-specific baseline hazards: ~ cause  + s(tend, by = cause)
#   - cause-specific effect of exposure: ~ group:cause
abortion_pam <- pamm(
  formula  = ped_status ~ cause  + s(tend, by = cause) + group:cause,
  data     = abortion_ped,
  method   = "fREML",
  engine   = "bam",
  discrete = TRUE)

plot(abortion_pam, page = 1)
summary(abortion_pam)
tidy_fixed(abortion_pam)
```


4. Visualize the CIFs separately for the coumarin exposure groups and for the
causes ‘spontaneous abortion’ and ‘induced abortion’.


```{r, echo = solution, eval = solution}
# create new data set for CIF prediction for each covariate specification of interest
# here we want predictions for
# - all time points
# - both causes that indicate abortion (life birth not of interest here)
# Important!!!:
# - Data has to be grouped by each separate covariate specifications
# - Data must be arranged properly (otherwise integration/cumulation will yield results)
cif_ndf <- abortion_ped %>%
  make_newdata(
    tend = unique(tend),
    cause = unique(cause)[c(1,3)],
    group = unique(group)
  ) %>%
  group_by(cause, group) %>%  # !!! important
  arrange(tend, cause, group) %>%
  add_cif(abortion_pam)

ggplot(cif_ndf, aes(x = tend, y = cif)) +
  geom_line(aes(col = cause)) +
  facet_wrap(~group) +
  ylim(c(0, 1))

# alternatively:
ggplot(cif_ndf, aes(x = tend, y = cif, fill = cause)) +
  geom_area(position = "stack") +
  facet_wrap(~group) +
  ylim(c(0, 1))
```
