---
title: "IWSM 2023 - short course"
subtitle: "Exercise I: PAMM basics"
author: "Andreas Bender, Fabian Scheipl"
---

```{r, include = FALSE, echo = FALSE}
solution <- TRUE#!!
options(width = 120)
library(knitr)
opts_chunk$set(
  fig.align = "center",
  echo  = solution,
  eval = solution
)
```



```{r, echo = TRUE, eval = TRUE, message = FALSE}
library(pammtools)
library(mgcv)
library(ggplot2)
theme_set(theme_bw())
library(survival)
```

The **`pammtools`** package contains the data set `tumor` which contains survival times of patients with stomach area tumors as well as as set of covariates:

- `days`, `status`: survival time and status
- `charlson_score`: comorbidity score (the higher the worse)
- `metastases`: ...
- ...

An overview of the data is given below

```{r, echo = TRUE, eval = TRUE}
head(tumor)
summary(tumor)
```


2. Familiarize yourself with the `as_ped` function. Use it to transform the `tumor`data into the "Piece-wise exponential data" (PED) format.

```{r}
tumor_ped <- tumor |> as_ped(Surv(days, status) ~ .)
head(tumor_ped)
```


3. Fit a PAMM (using package **`mgcv`**) that includes all covariates (as linear effects). Look at the summary of the model. Use functions `tidy_fixed` and `gg_fixed` to print and visualize the fixed effects estimates. Interpret the results.
Visualize the estimated (log-)baseline hazard.

```{r}
tumor_pam1 <- gam(
  formula = ped_status ~ s(tend, k = 20) + charlson_score + age + sex + transfusion +
    complications + metastases + resection,
  data    = tumor_ped,
  family = "poisson", # don't forget !!!
  offset  = offset, # don't forget !!!
  method  = "REML")

# alternatively
tumor_pam1 <- pamm(
  formula = ped_status ~ s(tend) + charlson_score + age + sex + transfusion +
    complications + metastases + resection,
  data = tumor_ped,
  method = "REML")

# better yet (faster)
tumor_pam1 <- pamm(
  formula = ped_status ~ s(tend) + charlson_score + age + sex + transfusion +
    complications + metastases + resection,
  data     = tumor_ped,
  method   = "fREML",
  engine   = "bam",
  discrete = TRUE)

# model summary
summary(tumor_pam1)
# extract coefs + CI's
tidy_fixed(tumor_pam1)
# forest plot
gg_fixed(tumor_pam1)
# Hazard Ratios: exp(beta)
tidy_fixed(tumor_pam1) |>
  mutate(across(where(is.numeric), exp))
# log baseline hazard
plot(tumor_pam1, las = 1)
```


4. Fit a Cox model to the same data, compare the estimated covariate effects to the ones obtained from the PAMM fit.

```{r}
tumor_cox1 <- coxph(Surv(days, status) ~ ., data = tumor)
tidy_fixed(tumor_cox1)
tidy_fixed(tumor_pam1)
cbind(Cox = coef(tumor_cox1), PAMM = coef(tumor_pam1)[2:8])
```


5. Use the `make_newdata` and `add_hazard` function to calculate the hazard ratio of a subject with complications, age = 53 vs. a subject without metastases and age = 47 (everything else being equal)

```{r }
tumor_ped |> make_newdata(complications = c("yes"), age = c(53)) |>
  add_hazard(tumor_pam1, ref = list(complications = c("no"), age = c(47))) |>
  select(hazard, ci_lower, ci_upper)
```


5. Run the `cox.zph` function to check for potential non-proportionality of hazards. What do you conclude?

```{r}
tumor_zph_cox1 <- cox.zph(tumor_cox1)
tumor_zph_cox1
plot(tumor_zph_cox1, var = 5)
```


6. We want to adress the problem with potential non-proportionality of `complications` and `metastases` variables. Create a new variable that represents the interaction of `complications` and `metastases` with four levels (i.e. all combinations of metastases yes/no, complications yes/no). Update your previous PAMM model to have (potentially) different baseline hazards for each of these combinations (i.e. stratification w.r.t. the new variable).


```{r}
tumor_ped <- tumor_ped |>
  mutate(strata = interaction(complications, metastases))
tumor_pam2 <- update(
  tumor_pam1,
  . ~ . - s(tend) - metastases - complications + strata +s(tend, by = strata),
  data = tumor_ped)
summary(tumor_pam2)
```


7. Use the `make_newdata` and `add_hazard` functions to create a data set that contains the estimated hazard for each interval/time-point for each combination of metastases and complications. Use this data set to visualize the four estimated baseline hazards.

```{r}
ndf_tumor <- tumor_ped |>
  make_newdata(
    tend        = unique(tend),
    strata      = unique(strata),
    age         = c(0), sex = c("male"),
    transfusion = c("no"),
    resection   = c("no")) |>
  add_hazard(tumor_pam2)

ggplot(ndf_tumor, aes(x = tend, y = hazard)) +
  geom_stephazard(aes(col = strata)) +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = strata), alpha = .1)
```

8. Now calculate the survival probabilities for all 4 strata using
  `add_surv_prob`, assuming an average person w.r.t. other covariates.

```{r}
ndf_tumor2 <- tumor_ped |>
  make_newdata(# sets other covariates to average/modus values by default
    tend        = unique(tend),
    strata      = unique(strata)) |>
  group_by(strata) |> # !!!! important !!!!
  add_surv_prob(tumor_pam2, overwrite = TRUE)
ggplot(ndf_tumor2, aes(x = tend, y = surv_prob)) +
  geom_surv(aes(col = strata)) +
  ylim(c(0, 1))
```


9. Calculate the time-dependent hazard ratio of a subject with complications and with metastases to a subject without complications with metastases (including confidence intervals) and visualize this hazard ratio over time

```{r}
ndf_hr <- tumor_ped |>
  make_newdata(tend = unique(tend), strata = c("yes.no"))|>
  add_hazard(tumor_pam2, reference = list(strata = c("yes.yes")))
ggplot(ndf_hr, aes(x = tend, y = hazard)) +
  geom_stephazard() +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = .3)
```
