---
title: "Survival Analysis W22/23"
subtitle: "Piecewise Exponential (Additive) Model"
author: "Andreas Bender"
---


```{r, echo = FALSE}
options(width = 120)
library(knitr)
opts_chunk$set(
  fig.align = "center"
)
```



```{r, message = FALSE}
library(dplyr)
library(purrr)
library(survival)
library(pammtools)
library(mgcv)
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
```

In this Notebook, we look at the Piecewise Exponential Model (PEM) and the Piecewise exponential Additive Model (PAM). Conceptually, the models are equivalent to the Cox(-PH) model:

$$h(t|\mathbf{x}) = h_0(t)\exp(\mathbf{x}^\top\boldsymbol{\beta})$$

This means, covariates affect the hazard multiplicatively and can be interpreted via hazard ratios as in the Cox-PH model.

The baseline hazard however, is estimated (semi-)parametrically via optimization of the (penalized) Poisson likelihood rather than non-parametrically as in the Cox model.


## Example Veteran' Data
First we use the `veteran` data to illustrate that the PEM is equivalent to the Cox-PH model (w.r.t. to $\boldsymbol{\beta}$ estimates) when there are no ties in the data.

```{r}
data("cancer", package = "survival")
head(veteran)
```

## Piecewise Exponential Model (PEM)
Below we fit the PEM model after removing duplicate times from the data and show that the estimate for age is equivalent to the results of `survival::coxph`.
Note that the baseline hazard is estimated via the `interval` variable, which is a factor variable and will be reference coded.


First we transform the data to the PED format:

```{r, fig.width = 8, fig.height = 4, out.width = "600px"}
veteran <- veteran %>%
  mutate(
    trt   = 1L * (trt == 2),
    prior = 1L * (prior == 10),
    status = ifelse(time > 400, 0, status), # censor obs after t = 400
    time = ifelse(time > 400, 400, time))
# remove ties to illustrate equivalence with Cox approach
vetu <- filter(veteran, !duplicated(time))
ped_vetu <- vetu %>%
  as_ped(Surv(time, status)~., cut = unique(vetu$time), id = "id")
ped_vetu %>% select(-diagtime) %>% slice(1:6)
```

Now we can fit the PEM to the transformed data:


```{r}
pem_baseline <- glm(
  formula = ped_status ~ interval - 1, # -1 so that each beta_j directly corresponds to log(h_j)
  data    = ped_vetu,
  family  = poisson(),
  offset  = offset
)
coef(pem_baseline)[1:10] # beta0 = log-hazard in interval j=1, beta1 = log(h_2)
exp(coef(pem_baseline))[1:10]
pem_smry <- ped_vetu %>%
  group_by(interval) %>%
  summarize(
    t_j        = unique(tend),
    len_j      = unique(tend - tstart),
    d_j        = sum(ped_status),
    n_j        = n(),
    o_j        = sum(exp(offset)),
    rel_freq_j = d_j/n_j) %>%
  mutate(
    h_j = exp(coef(pem_baseline)),
    H_j = cumsum(h_j * len_j))
pem_smry |> head()# relative frequency of events in interval j = hazard in interval j (if len_j = 1)

## relationship to Nelson-Aalen estimator
cph_baseline <- coxph(Surv(time, status)~1, data = vetu)
H_cox <- basehaz(cph_baseline)
pem_smry %>% cbind(H_cox) %>% head()
```


Include `karno` variable:

$$\begin{align}
h(t|\mathbf{x})
  & = h_0(t)\exp(\beta_1 x_{karno})\\
  & = \exp(\beta_{0j} + \beta_1x_{karno})\ \forall t \in (a_{j-1}, a_j]
\end{align}$$

```{r}
pem_karno <- glm(
  formula = ped_status ~ interval + karno,
  data    = ped_vetu,
  family  = poisson(),
  offset  = offset)
coef(pem_karno)[1:10]
```

Thus, the log baseline hazard is

  - $\beta_{01}$ = `(Intercept)` = `r round(coef(pem_karno)[1], 2)` in interval $(0,1]$
  - $\beta_{02}$ = `sum(coef(pem_karno)[c(1,2)])` = `r round(sum(coef(pem_karno)[1:2]), 2)` in interval $(1,2]$
  - ...
  - $\beta_{05}$ = `sum(coef(pem_karno)[c(1,5)])` = `r round(sum(coef(pem_karno)[c(1,5)]),2)` in interval $(4, 7]$
  - ...


$\beta_{1}$ estimates are the same for the PEM and the Cox-PH model:

```{r}
## cox model for comparison
cph_karno <- coxph(Surv(time, status) ~ karno, data = vetu)
## compare coefficients
cbind(
  pem = coef(pem_karno)["karno"],
  cox = coef(cph_karno))
```

### Piecewise exponential Additive Model (PAM)

In the PAM, the baseline hazard is estimated differently (basis function representation + penalized estimation of basis coefficients):

$$\begin{align}
h(t|\mathbf{x})
  & = h_0(t)\exp(\beta_1 x_{karno})\\
  & = \exp(\underbrace{\beta_0 + f_0(t)}_{g_0(t)} + \beta_1x_{karno})
\end{align}$$

with $f_0(t) = \sum_{\ell=1}^L \gamma_{\ell}B_\ell(t_j)\ \forall t \in (a_{j-1},a_j]$


```{r}
## compare to PAM
pam_karno <- gam(
  formula = ped_status ~ s(tend) + karno,
  data    = ped_vetu,
  family  = "poisson",
  offset  = offset)
```

```{r}
base_pem  <- ped_vetu %>% ped_info() %>% mutate(karno = 0) %>%
  add_hazard(pem_karno)
base_pam <- ped_vetu %>% ped_info() %>% mutate(karno = 0) %>%
  add_hazard(pam_karno)
base_pam %>% select(tend, interval, karno, hazard, ci_lower, ci_upper)
ggplot(base_pam, aes(x = tend, y = hazard)) +
  geom_stephazard() +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = .3) +
  geom_stephazard(data = base_pem, col = 2)
```

Note that the (absolut value of the) baseline hazard doesn't necessarily has to have a usefull interpretation, as it is the hazard for $\mathbf{x} = \mathbf{0}$.

The estimate of $\beta_{x_{karno}}$ also necessarily differs between the PEM and PAM, but usually not by much:

```{r}
cbind(
    pem = coef(pem_karno)["karno"],
    pam = coef(pam_karno)["karno"],
    cox = coef(cph_karno)["karno"])
```

<br><br><br>

More covariates:

```{r}
mod_list <- list(pem = pem_karno, cph = cph_karno, pam = pam_karno) %>%
  imap(~update(.x, .~. + trt + prior + karno))

# summary output PAM
summary(mod_list[["pam"]])
# compare coefficient estimates
mod_list %>% map_dfr(~coef(.x)[c("trt", "prior", "karno")], .id = "model")
```


## Time-varying effects

- In the PAM, time (e.g., `tend`) is simply another covariate.
- So far we used it to estimate the (log) baseline hazard $h_0(t) = \exp(g_0(t))$.
- Time-varying effects, are simply interactions of the covariate time (`tend`) and other covariates (similar to the extended Cox-PH model with the data in start-stop format, see `cox-non-ph.Rmd`)
- Interactions of time with categorical covariates are equivalent to "stratification" in Cox-PH models
- The type of time-variation is dependent on the specification of the interaction:
  + $f(t)x$ non-linearly time-varying effect, linear in $x$
  + $f(t, x)$ non-linear, non-linearly time-varying effect
- $f$ can either be an assumed parametric function, e.g. $f=\log$, or can be estimated from the data


### Example Veteran's Data

#### Linear, non-linearly time-varying effect (known $f$)
Let's consider once again the estimation of the time-varying effect of the `karno` variable (see `W07-3-cox-non-ph.Rmd`):

```{r}
cph_tt <- coxph(
      formula = Surv(time, status) ~ trt + prior + karno + tt(karno),
      data    = veteran,
      tt      = function(x, t, ...) x * log(t + 20))
summary(cph_tt)
```

The equivalent PAM is given below:

```{r}
# create a covariate that is the assumed transformation of time
ped <- veteran %>% as_ped(Surv(time, status)~., id = "id") %>%
  mutate(logt20 = log(tstart + (tstart - tend) / 2 + 20))
# fit the model: pamm is a short cut for gam(..., family = poisson(), offset = offset)
pam_tt <- pamm(ped_status ~ s(tend) + trt + prior + karno + karno:logt20, data = ped)
cbind(
  pam = coef(pam_tt)[2:5],
  cox = coef(cph_tt))
```

Compare fits:
```{r}
# compare fits
t <- seq(0, 400, by = 10)
plot(x = t, y = coef(cph_tt)["karno"] + coef(cph_tt)["tt(karno)"] * log(t + 20),
  type = "l", ylab = "Beta(t) for karno", ylim = c(-.1, .05), las = 1,
  col = 1)
lines(
  x    = t,
  y    = coef(pam_tt)["karno"] + coef(pam_tt)["karno:logt20"] * log(t + 20),
  col  = 2, lty = 2)
```

#### Linear, non-linearly time-varying effect (unknown $f$)
However, we don't have to assume a parametric shape of the time variation and instead estimate it from the data:

$$\begin{align}
h(t|\mathbf{x})
  & = h_0(t)\exp(\beta_1x_{trt} + \beta_2 x_{prior} + f_{karno}(t)x_{karno})\\
  & = \exp(\underbrace{\beta_0 + f_0(t)}_{g_0(t)} + \beta_1x_{trt} + \beta_2 x_{prior} + f_{karno}(t)x_{karno})
\end{align}$$

with $f_{karno}(t) = \sum_{m=1}^M \gamma_{m}B_m(t_j)\ \forall t \in (a_{j-1},a_j]$

```{r}
pam_tt2 <- pamm(ped_status ~ s(tend) + trt + prior + s(tend, by = karno), data = ped)
summary(pam_tt2)
```

Comparison to parametric form:

```{r}
plot(pam_tt2, select = 2, ylim = c(-.1, .1))
lines(
  x    = t,
  y    = coef(pam_tt)["karno"] + coef(pam_tt)["karno:logt20"] * log(t + 20),
  col  = 2, lty = 2)
abline(h = 0, lty = 3)
```

- `karno` has hazard-decreasing effect in the beginning
- the effect tends to 0 towards end of follow-up


#### Non-Linear, non-linearly time-varying effect (unknown $f$)

In the previous models we assumed that the effect of `karno` is non-linearly time-varying, but linear in `karno`. We can relax this assumption by estimating a bivariate function $f(t, x_{karno})$, e.g. using a tensor product smooth:

$$\begin{align}
h(t|\mathbf{x})
  & = h_0(t)\exp(\beta_1x_{trt} + \beta_2 x_{prior} + f(t, x_{karno}))\\
  & = \exp(\beta_0 + f_0(t) + \beta_1x_{trt} + \beta_2 x_{prior} + f(t, x_{karno}))
\end{align}$$

with $f(t, x_{karno}) = \sum_{m=1}^M \sum_{r=1}^R \gamma_{m,r}B_m(t_j)B_{r}(x_{karno})\ \forall t \in (a_{j-1},a_j]$

```{r}
pam_te <- pamm(ped_status ~ s(tend, k = 20) + trt + prior + te(tend, karno), data = ped)
summary(pam_te)
```

```{r, fig.width = 9, fig.height = 3}
gg_tensor(pam_te, ci = TRUE)
```

```{r, fig.width = 9, fig.height = 3}
p_tensor <- gg_tensor(pam_te) +
  geom_vline(xintercept = c(1, 100, 200), lty = 3) +
  geom_hline(yintercept = c(40, 75, 95), lty = 3)
# slices through the surface for fixed karno values
p_slice_karno <- gg_slice(ped, pam_te, "karno", tend = unique(tend), karno = c(40, 75, 95)) +
  geom_hline(yintercept = 0, lty = 2)
# slices through the surface for fixed times
p_slice_time <- gg_slice(ped, pam_te, "karno", tend = c(1, 100, 200), karno = seq_range(karno, 50)) +
  geom_hline(yintercept = 0, lty = 2)
p_tensor + p_slice_karno + p_slice_time
```


### Example Liver Resection

```{r}
data("tumor", package = "pammtools")
head(tumor)
ped_tumor <- tumor %>%
  as_ped(Surv(days, status) ~.) %>%
  mutate(strata = interaction(complications, metastases))
```

We use an interaction term between time and a categorical covariate to estimate
group-specific baseline hazards:

$$\begin{align}
h(t|\mathbf{x})
  & = h_0(t,k)\exp(\mathbf{x}^\top\boldsymbol{\beta}))\\
  & = \exp(\beta_{0,k} + f_{0,k}(t) + \mathbf{x}^\top\boldsymbol{\beta})
\end{align}$$

with $k=1,\ldots,K$ and $K$ number of groups.

```{r}
pam_strata <- pamm(
  formula  = ped_status ~ strata + s(tend, by = strata, k = 20) + charlson_score + transfusion + sex + resection,
  data     = ped_tumor,
  engine   = "bam",
  method   = "fREML",
  discrete = TRUE)
summary(pam_strata)
```


Visualize the estimated hazards in the 4 groups:

```{r}
ndf <- ped_tumor %>%
  make_newdata(tend = unique(tend), strata = unique(strata)) %>%
  add_hazard(pam_strata)
ggplot(ndf, aes(x = tend, y = hazard, group = strata)) +
  geom_stephazard(aes(col = strata)) +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = strata), alpha = .3)
```

We can also visualize the differences of log-hazard of two groups with confidence intervals for the differences:

```{r}
gg_partial(ped_tumor, pam_strata, term = "strata", tend = unique(tend),  strata = c("yes.no"),
  reference = list(strata = "yes.yes")) +
  geom_hline(yintercept = 0, lty = 2)
gg_partial(ped_tumor, pam_strata, term = "strata", tend = unique(tend),  strata = c("yes.no"),
  reference = list(strata = "no.no")) +
  geom_hline(yintercept = 0, lty = 2)
```

Visualize survival probabilities in the 4 groups (other covariates set to mean/modus):

```{r}
ndf <- ndf %>%
  group_by(strata) %>% # important!!
  add_surv_prob(pam_strata)
ggplot(ndf, aes(x = tend, y = surv_prob, group = strata)) +
  geom_surv(aes(col = strata)) +
  geom_ribbon(aes(ymin = surv_lower, ymax = surv_upper, fill = strata), alpha = .3)
```


## Left truncation

### Example Infant Mortality
```{r}
data("infants", package = "eha")
infants %>% slice(97, 100)
```

Data transformation to PED format:

```{r}
ped_lt <- infants %>% as_ped(Surv(enter, exit, event)~.)
ped_lt %>%
  filter(id %in% c(97, 100)) %>%
  select(id, interval, ped_status, mother, sex)
pam_lt <- pamm(
  formula = ped_status ~ s(tend, k = 20) + mother + age + sex + parish + civst + ses,
  data = ped_lt)
summary(pam_lt)
# comparison to Weibull AFT and CoxPH
wb_lt <- eha::phreg(
  formula = Surv(enter, exit, event) ~ mother + age + sex + parish + civst + ses,
  data    = infants)
cph_lt <- coxph(
  formula = Surv(enter, exit, event) ~ mother + age + sex + parish + civst + ses,
  data    = infants)
cbind(
  cph = exp(coef(cph_lt)),
  pam = exp(coef(pam_lt)[2:7]),
  wb  = exp(coef(wb_lt)[1:6]))
```


#### Checking independence assumption of left truncation times

```{r}
ped_lt <- ped_lt %>%
  group_by(id) %>%
  mutate(entry_time = tstart[1]) # here equivalent to childs age at entry

pam_lt2 <- update(pam_lt, .~. + s(entry_time), data = ped_lt)
summary(pam_lt2)
cbind(
  cph  = exp(coef(cph_lt)),
  pam  = exp(coef(pam_lt)[2:7]),
  pam2 = exp(coef(pam_lt2)[2:7]),
  wb   = exp(coef(wb_lt)[1:6]))
```
