---
title: "Survival Analysis W22/23"
subtitle: "Frailty Models"
author: "Andreas Bender"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 4,
  fig.heiht = 4,
  fig.align = "center"
)
```


```{r, message = FALSE}
library(pammtools)
library(survival)
```


## Multi-center Study

- (subset) of critically ill patients admited to ICU (follow-up of 30 days)
- various covariates
  + sex (`Gender`)
  + age (`Age`)
  + admission category (`AdmCatID`)
  + Appache II Score (`ApacheIIScore`)
  + BMI (`BMI`)
  + Diagnosis (`DiagID2`)
- data collected in a multi-center study (multiple hospitals/centers from different countries): `CombinedicuID`
- event of interest: Time until death in hospital (discharge considered a censoring event): `survhosp`, `PatientDied`

```{r}
data(patient, package = "pammtools")
head(patient)
levels(patient$AdmCatID)
levels(patient$DiagID2)
length(unique(patient$CombinedicuID))# -> many levels
```

## Cox Frailty Model

### Normal distribution

Here we assume log-Gaussian frailty:

```{r}
cph_gauss <- coxph(
  formula = Surv(survhosp, PatientDied) ~ Gender + Age + AdmCatID +
    ApacheIIScore + BMI + DiagID2 +
    frailty(CombinedicuID, distribution = "gaussian", method = "aic"),
    data = patient)
summary(cph_gauss)
```

We can check the assumption of $b_i \sim N(0, \sigma_b^2)$ by plotting the empirical distribution of (predicted) random effects vs. theoretical quantiles of the Gaussian:

```{r, fig.width = 4, fig.height = 4, out.width = "300px"}
frail_gauss <- cph_gauss$frail
qqnorm(frail_gauss, las = 1)
qqline(frail_gauss, col = 2, lwd = 2)
```

### Gamma distribution

Here we assume Gamma frailty:

```{r}
cph_gamma <- coxph(
  formula = Surv(survhosp, PatientDied) ~ Gender + Age + AdmCatID +
    ApacheIIScore + BMI + DiagID2 +
    frailty(CombinedicuID, method = "aic"),
  data = patient)
summary(cph_gamma)
cbind(gauss = coef(cph_gauss), gamma = coef(cph_gamma))
```

Comparison of Gaussian vs. Gamma
```{r}
frail_gamma <- cph_gamma$frail
plot(frail_gauss ~ frail_gamma, las = 1)
abline(0, 1, col = 2)
```


## PAMM

```{r, fig.width = 4, fig.height = 4}
ped <- patient |> as_ped(Surv(survhosp, PatientDied) ~ .)
pamm <- mgcv::bam(
  formula =  ped_status ~ s(tend) + Gender + Age + AdmCatID + ApacheIIScore +
    BMI + DiagID2 + s(CombinedicuID, bs = "re"),
  data = ped,
  family = poisson(),
  offset = offset,
  discrete = TRUE)
summary(pamm)
# comparison to cph fixed effect estimates
cbind(pamm = coef(pamm)[2:15], cph_gauss = coef(cph_gauss))
# visualization (qqplot)
gg_re(pamm)
```

```{r fig.width = 4, fig.height = 4}
# comparison of random effects
plot(tidy_re(pamm)$fit ~ frail_gauss, las = 1)
abline(0, 1, col = 2)
```
