---
title: "Survival Analysis W22/23"
subtitle: "Recurrent Events"
author: "Andreas Bender"
---

```{r, message = FALSE}
library(survival)
library(pammtools)
library(purrr)
library(dplyr)
# visualization
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
```

```{r, message = FALSE, echo = FALSE, eval = TRUE}
# references
library(RefManageR)
BibOptions(check.entries = FALSE, hyperlink=TRUE, style = "markdown",
  max.names = 1)
bib <- ReadBib("../references.bib", check = FALSE)
options(width = 120)
```



## Chronic Granulotamous Disease (CGD)

- RCT for effect of interferon gamma (treatment) on patients with CGD
- CGD is an inherited disorder that manifests in recurrent infections (with skin leasons)

```{r}
data("cgd", package = "survival")
cgd %>% select(-random, -hos.cat, -center) %>% head()
table(cgd$enum)
```

```{r}
cgd <- cgd %>%
  mutate(
    steroids = as.factor(ifelse(steroids, "yes", "no")),
    gap = tstop - tstart,
    enum2 = ifelse(enum > 3, "4+", enum))
table(cgd$enum2)
# assumption of renewal process without subject specific correlation or dependence on event
cph1 <- coxph(
  formula = Surv(gap, status) ~ treat + steroids + age + inherit,
  data = cgd)
summary(cph1)
cph2 <- update(cph1, .~. + frailty(id))
cph3 <- update(cph1, .~. + strata(enum2))
cph4 <- update(cph2, .~. + strata(enum2))
```

### Treatment effect
```{r}
imap_dfr(
  .x = list(cph1, cph2, cph3, cph4),
  .f = ~ {
    coef_tab <- summary(.x)$coefficients
    colnames(coef_tab)[ncol(coef_tab)] <- "p"
    coef_tab[1, c("coef", "se(coef)", "p")]
  },
  .id = "model")
```

### Frailties
```{r}
summary(cph2)$coefficients[5,]
summary(cph4)$coefficients[5,]
summary(cph2)$print2
summary(cph4)$print2
```

### Strata
```{r, fig.width = 8, fig.height = 3}
bh3 <- basehaz(cph3)
bh4 <- basehaz(cph4)
p3 <- ggplot(bh3, aes(x = time, y = hazard, col=strata))+ geom_stephazard() +
  ylab("cumulative hazard")
p4 <- p3 %+% bh4
p3 + p4
```

$\rightarrow$ baseline hazard almost identical for both analyses <br>

$\rightarrow$ higher baseline hazard for increased event number <br>

$\rightarrow$ when stratification is not included, the correlation induced by "event number" is (partially) absorbed by the random effects (hence the differences in random effect variance in the two analyses)

<!--
# References

```{r, results = "asis", warning = FALSE, echo = FALSE}
PrintBibliography(bib, start = 1, end = 2)
```
 -->
