---
title: "Survival Analysis W22/23"
subtitle: "Recurrent Events"
author: "Andreas Bender"
---

```{r, message = FALSE}
library(survival)
library(pammtools)
library(mgcv)
library(purrr)
library(dplyr)
# visualization
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
```

```{r, echo = FALSE, message = FALSE}
# references
library(RefManageR)
BibOptions(check.entries = FALSE, hyperlink=TRUE, style = "markdown",
  max.names = 1)
bib <- ReadBib("../references.bib", check = FALSE)
```



## Childhood Malaria

- A randomize controled trial following a birth cohort for one year in Uganda `r Citep(bib, "kakuru_impact_2020")`
- Goal: study of two intermittent preventive treatments on incidence of childhood malaria

```{r}
malaria <- readRDS("../data/malaria-recurrence.Rds") |>
  mutate(Txarm = factor(Txarm, levels = c(0,1), labels = c("no", "yes"))) |>
  mutate(id = as.factor(id))
table(malaria$enum)
malaria |> select(-date, -enddate) |>
  filter(id %in% c("31161", "32260", "31088"))
```

Here we aim to analyze the data using PAMMs on the calendar time scale, additionally estimating the seasonality of occurrences

### Transformation to PED format

- First we transform the data to the PED format
- For recurrent events, we have to provide the name of the column that stores the informaiton about the event number via the `transition` argument
- Because we want to work on the calendar time scale, the `Surv` object must be in start-stop format
- additionally we have to specify `timescale = "calendar"`, as `"gap"` is the default

```{r}
ped <- malaria |>
  as_ped(
    formula = Surv(start, time, status)~Txarm + gender + preterm + Gravidity + dobday,
    id = "id",
    transition = "enum",
    timescale = "calendar")
# create day of year variable for seasonality effects
ped$doy <- ped$dobday + ped$tstart
ped$doy = ifelse(
  ped$doy>730,
  ped$doy-730,
  ifelse(ped$doy>365,ped$doy-365,ped$doy))


ped |>
  filter(id %in% c("31161", "32260", "31088")) |>
  group_by(id, enum) |>
  slice(1, n()) |>
  knitr::kable()
```


### Fit the model

```{r}
pamm_malaria <- pamm(
  formula  = ped_status ~ s(tend) + gender * Txarm + preterm + Gravidity + s(id, bs = "re"),
  data     = ped,
  engine   = "bam",
  method   = "fREML",
  discrete = TRUE)
summary(pamm_malaria)
```

### Model with seasonality

```{r}
pamm_malaria2 <- update(pamm_malaria, .~. + s(doy, bs = "cc"))
summary(pamm_malaria2)
```

### Visualise seasonality effect

```{r}
ndf <- ped |> make_newdata(doy = 1:365) |>
  mutate(date = as.Date("2017-01-01") + (doy-1)) |>
  add_term(pamm_malaria2, term = "doy")
head(ndf)
ggplot(ndf, aes(x = date, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = .3) +
   scale_x_date(expand = c(0, 0),date_breaks = "1 month",date_labels = "%b")
```


### Interaction of child age with seasonality

```{r, fig.width = 9, fig.height = 3}
pamm_malaria3 <- update(
  pamm_malaria,
  .~. -s(tend) + te(tend, doy, bs = c("cr", "cc")))
summary(pamm_malaria3)
gg_tensor(pamm_malaria3, ci = TRUE) +
   xlab("Age") + ylab("Season")
```

```{r fig.width = 6, fig.height = 4}
slice_age <- gg_slice(ped, pamm_malaria3, term = "doy", tend  = c(50, 150,  300), doy=1:365)
slice_doy <- gg_slice(ped, pamm_malaria3, term = "doy", tend  = unique(tend), doy=c(50, 150, 300)) + xlab("age")
slice_age + slice_doy
```

$\rightarrow$ indicates that no interaction necessary

# References

```{r, results = "asis", warning = FALSE, echo = FALSE}
PrintBibliography(bib, start = 1, end = 2)
```
