---
title: "Survival Analysis W22/23"
subtitle: "Competing Risks"
author: "Andreas Bender"
---

```{r, message = FALSE}
library(pammtools)
library(purrr)
library(dplyr)
# visualization
library(ggplot2)
theme_set(theme_bw())
```

```{r, message = FALSE, echo = FALSE, eval = TRUE}
# references
library(RefManageR)
BibOptions(check.entries = FALSE, hyperlink=TRUE, style = "markdown",
  max.names = 1)
bib <- ReadBib("../references.bib", check = FALSE)
options(width = 120)
library(knitr)
opts_chunk$set(
  dev        = "png",
  dpi        = 500,
  fig.align  = "center",
  out.width  = "400px",
  fig.width  = 5,
  fig.height = 5,
  collapse   = TRUE
  )
```



## ICU patients (`sir.adm`)

The example below follows Beyersmann et al. (2008): "Competing Risks and Multistate Models with R"

```{r}
data(sir.adm, package = "mvna")
sir.adm %>% filter(id %in% c(41, 395, 17058))
table(sir.adm$status)
sum(sir.adm$status != 0)
```

- 747 ICU patients
- endpoints:
  + discharge from the ICU (`status` = 1): competing event
  + death (`status` = 2): event of interest
- variable of interest  `pneu`: A patient's pneumonia status at admission (0 = no pneumonia, 1 = pneumonia)


## PAM

### Data in PED format
```{r}
cut <- unique(sir.adm$time[sir.adm$status != 0])
ped_list <- as_ped(sir.adm, Surv(time, status)~., combine = FALSE, cut = cut)
str(ped_list,1)
ped_list[[1]] %>%
  filter(id %in% c(41, 395, 17058)) %>%
  group_by(id) %>% slice(1, n())
ped_list[[2]] %>%
  filter(id %in% c(41, 395, 17058)) %>%
  group_by(id) %>% slice(1, n())
```

Note that the total number of events between the two data sets doesn't change
```{r}
sum(ped_list[[1]]$ped_status, ped_list[[2]]$ped_status)
```

### Cause-specific hazards (separate models)

```{r}
pam_list <- map(ped_list, ~pamm(ped_status ~ s(tend) + pneu, data = .x))
map(pam_list, ~summary(.x))
map_dfc(pam_list, ~round(exp(coef(.x)["pneu"]), 2))
```

### Cause-specific hazards (one model)

```{r}
# combined = TRUE is the default
ped <- as_ped(sir.adm, Surv(time, status)~., combine = TRUE) %>%
  mutate(cause = as.factor(cause))
dim(ped)
str(ped)
ped %>%
  filter(id %in% c(41, 395, 17058)) %>%
  group_by(id, cause) %>%
  slice(1, n())

sum(ped$ped_status)

# Fit the model
pam <- pamm(ped_status ~ s(tend, by = cause) + cause*pneu, data = ped)
# what would it mean if we fit
# pamm(ped_status ~ s(tend) + cause* pneu, data = ped)
summary(pam)
round(cbind(Discharge = exp(coef(pam)[3]), Death = exp(sum(coef(pam)[3:4]) )), 2)
```

### Cumulative Incidence Functions

- pneumonia doesn't appear to have an effect on the death hazard
- this does *not* imply that pneumonia has no effect on mortality

```{r, fig.width = 8, fig.height = 3, out.width = "600px"}
ndf <- ped %>% make_newdata(
  tend  = unique(tend),
  pneu  = unique(pneu),
  cause = unique(cause))
ndf %>% group_by(cause) %>% slice(1:3)
ndf <- ndf %>%
  group_by(cause, pneu) %>% # important!
  add_cif(pam) %>%
  ungroup() %>%
  mutate(
    cause = factor(cause, labels = c("Discharge", "Death")),
    pneu  = factor(pneu, labels = c("No Pneumonia", "Pneumonia"))
  )
# visualization
ggplot(ndf, aes(x = tend, y = cif)) +
  geom_line(aes(col = pneu)) +
  geom_ribbon(aes(ymin = cif_lower, ymax = cif_upper, fill = pneu), alpha = .3) +
  facet_wrap(~cause) +
  labs(
    y     = "CIF",
    x     = "time",
    color = "Pneumonia",
    fill  = "Pneumonia"
  )
```

- Pneumonia reduces the hazard for Discharge $\rightarrow$ subjects admitted with pneumonia stay on ICU longer
- While they are in the ICU, the are exposed to an unchanged hazard for death
- As a consequence more patients with pneumonia die than patients without pneumonia
